---
title: "Epimutations in INMA Sabadell"
author: "Carlos Ruiz"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Introduction

In this document, we will summarize the main results of the epimutations detection in INMA Sabadell samples measured at 0, 4 and 8 years.


```{r}
library(tidyverse)
library(cowplot)
library(minfi)
library(robustbase)
library(epimutacions)
library(UpSetR)
```

# Normalized data

First, we will compare the results obtained computing the epimutations in the dataset without applying the residuals.


```{r}
## 0 years
load("../INMA_comb.normalizedComBat.autosomic.filterAnnotatedProbes.withNA.GenomicRatioSet.Rdata")
gset.0 <- gset

## 4 years
load("../MeDALL_all.normalizedComBat.autosomic.filterAnnotatedProbes.withNA.GenomicRatioSet.Rdata")

gset.sab <- gset[, gset$Sample_Group == "Age_4" & grepl("^04|SAB", gset$Sample_Name)]
gset.4  <- gset.sab[, !duplicated(gset.sab$idnum)]

## 8 years
load("../results/preprocess/HELIX/HELIX.withNA.GenomicRatioSet.Rdata")
helix <- gset[, gset$cohort != "MOBA" & gset$h_ethnicity_3cat == "WhiteEur_WhiteOther"]
```

```{r}
make_res_df <- function(res.list, pheno, dataset){
  nMethod <- sapply(res.list, nrow)
  nMethod[sapply(nMethod, is.null)] <- 0
  methods <- names(res.list)
  names(methods) <- methods
  Reduce(rbind, res.list) %>%
    mutate(method = rep(methods, unlist(nMethod)),
           dataset = dataset) %>%
    left_join(pheno %>% 
                data.frame() %>% 
                select(Sample_Name, Sex, idnum) %>% 
                mutate(sample = Sample_Name,
                       idnum = as.character(idnum)), by = "sample")
}
make_sum_df <- function(res_df){
  res_df %>%
  group_by(method, sample, Sex, idnum, dataset) %>%
  summarize(n = sum(start != 0 & (is.na(pvalue) | pvalue < 0.05/40408))) %>%
  mutate(n_cat = ifelse(n > 1, "1+", as.character(n)))
}

```

```{r}
load("../INMA_comb.epimutations.allSamples.Rdata")
load("../INMA4.epimutations.allSamples.Rdata")
load("../results/epimutations/HELIX.epimutations.allSamples.Rdata")

helixp <- colData(helix)
helixp$idnum <- sapply(strsplit(helixp$SampleID, "_"), `[`, 2)
helixp$Sex <- helixp$e3_sex
helixp$Sample_Name <- helixp$SampleID

res.inma0.df <- make_res_df(res.inma0.list, pheno = colData(gset.0), dataset = "Cord Blood") 
inma0.sum.df <- make_sum_df(res.inma0.df)


res.inma4.df <- make_res_df(res.inma4.list, pheno = colData(gset.4), dataset = "4 years") 
inma4.sum.df <- make_sum_df(res.inma4.df)

res.helix.df <- make_res_df(res.helix.list, pheno = helixp, dataset = "8 years") 
helix.sum.df <- make_sum_df(res.helix.df)

all.sum.df <- rbind(inma0.sum.df, inma4.sum.df, helix.sum.df) %>%
  mutate(dataset = factor(dataset, levels = c("Cord Blood", "4 years", "8 years")))
methods <- unique(all.sum.df$method)
names(methods ) <- methods

all.res.df <- rbind(res.inma0.df, res.inma4.df, res.helix.df) %>%
  mutate(dataset = factor(dataset, levels = c("Cord Blood", "4 years", "8 years")))
```

```{r}
all.sum.df %>%
  ggplot(aes(x = n_cat)) +
  geom_bar() +
  theme_bw() +
  facet_grid(dataset ~ method, scales = "free_y")

sum.props <- lapply(methods, function(x) {
  
  sub <- subset(all.sum.df, method == x)
   round(prop.table(table(sub$dataset, sub$n_cat), margin = 1)*100, 2)

})
sum.props
```

We observed that more children had epimutations, and a higher proportion of children had more than 1 epimutation in cord blood than at 8 years. This tendency is common for the four methods, although the proportiongs are very different. Thus, in barbosa around 17% of individuals had an epimutations on cord blood, while less than 1% have an epimutation at 8 years. For beta, this discrepancy is higher for beta, passing from 35% to 2%. manova and mlm detected epimutations in almost all samples at cord blood (more than 85%), while they detected epimutations at half of ndividuals at 8 years (around 60%).

```{r}
sab.sum.df <- subset(all.sum.df, dataset != "8 years" | grepl("SAB", sample))

upsetList <- lapply(unique(sab.sum.df$dataset), function(x) {
  unique(subset(sab.sum.df, dataset == x)$idnum)
  })
names(upsetList) <- unique(sab.sum.df$dataset)
upset(fromList(upsetList))
```

110 individuals were shared between the three time points. Next, we have some overlap between cord blood and 4 years and between cord blood and 8 years.

```{r}
tab <- table(sab.sum.df$idnum)
dup <- names(tab)[tab > 4]
com.res.df <- subset(all.res.df, idnum %in% dup & chromosome != 0 & (is.na(pvalue) | pvalue < 0.05/40408))
long_sum <- com.res.df %>%
  group_by(idnum, method, epi_region_id) %>%
  summarize(n_dataset = length(unique(dataset))) %>%
  filter(n_dataset > 1) %>%
  arrange(method)

table(long_sum$method, long_sum$n_dataset)
```

6 epimutations were found in two timepoints with beta, while 29 and 46 epimutations were indentifed in two time-points and 1 in the three timepoints. With barbosa, all epimutations were only found in a timepoint. 

```{r}
epi_com_beta <- subset(com.res.df, idnum == "304" & method %in% c("beta", "barbosa"))

p1_cord <- plot_epimutations(epi_com_beta[1,], gset.0) +
  ggtitle("Cord Blood")
p1_4years <- epi_com_beta %>%
  mutate(sample = "04_304_4") %>%
  plot_epimutations(methy = gset.4) +
  ggtitle("4 years")
p1_8years <- epi_com_beta %>%
  mutate(sample = "SAB_304_1X",
         cpg_ids = "cg27182551,cg13431205,cg18481241,cg25676335,cg00059930,cg26740195") %>%
  plot_epimutations(methy = helix) +
  ggtitle("8 years")
plot_grid(p1_cord, p1_4years, p1_8years, nrow = 3)
```

This epimutation is observed at cord blood and 4 years, while at 8 years this individual still has the most extreme values. 


```{r}
epi_com_beta <- subset(com.res.df, idnum == "245" & method %in% c("beta", "barbosa"))

p1_cord <- plot_epimutations(epi_com_beta[1,], gset.0) +
  ggtitle("Cord Blood")
p1_4years <- epi_com_beta %>%
  mutate(sample = "04_245_4") %>%
  plot_epimutations(methy = gset.4) +
  ggtitle("4 years")
plot_grid(p1_cord, p1_4years, nrow = 2)
```



```{r}
epi_com_beta <- subset(com.res.df, idnum == "364" & method %in% c("beta", "barbosa"))

p1_cord <- plot_epimutations(epi_com_beta[1,], gset.0) +
  ggtitle("Cord Blood")
p1_4years <- epi_com_beta %>%
  mutate(sample = "04_364_4") %>%
  plot_epimutations(methy = gset.4) +
  ggtitle("4 years")
plot_grid(p1_cord, p1_4years, nrow = 2)
```

These two epimutations are observed at cord blood and 4 years. However, these individuals are not present at 8 years.


```{r}
epi_com_beta2 <- subset(com.res.df, idnum == "519" & method %in% c("beta", "barbosa"))

p2_cord <- plot_epimutations(epi_com_beta2[1,], gset.0) +
  ggtitle("Cord Blood")
p2_4years <- epi_com_beta2 %>%
  mutate(sample = "04_519_4") %>%
  plot_epimutations(methy = gset.4) +
  ggtitle("4 years")
p2_8years <- epi_com_beta2 %>%
  mutate(sample = "SAB_519_1X") %>%
  plot_epimutations(methy = helix) +
  ggtitle("8 years")
plot_grid(p2_cord, p2_4years, p2_8years, nrow = 3)
```

This epimutation is observed at cord blood and 4 years, while at 8 years the individual have value similar to the population.

```{r}
epi_com_mlm <- subset(com.res.df, idnum == "356" & method == "mlm" & epi_region_id == "chr8_145637525")

p3_cord <- plot_epimutations(epi_com_mlm[1,], gset.0) +
  ggtitle("Cord Blood")
p3_4years <- epi_com_mlm %>%
  mutate(sample = "04_356_4") %>%
  plot_epimutations(methy = gset.4) +
  ggtitle("4 years")
p3_8years <- epi_com_mlm %>%
  mutate(sample = "SAB_356_1X") %>%
  plot_epimutations(methy = helix) +
  ggtitle("8 years")
plot_grid(p3_cord, p3_4years, p3_8years, nrow = 3)
```

The epimutation detected by mlm at the three time-points is not very clear.

```{r}
epi_com_manova <- subset(com.res.df, idnum == "235" & method == "manova" & epi_region_id == "chr11_117314243")

p4_cord <- plot_epimutations(epi_com_manova[1,], gset.0) +
  ggtitle("Cord Blood")
p4_4years <- epi_com_manova %>%
  mutate(sample = "04_235_4") %>%
  plot_epimutations(methy = gset.4) +
  ggtitle("4 years")
p4_8years <- epi_com_manova %>%
  mutate(sample = "SAB_235_1X") %>%
  plot_epimutations(methy = helix) +
  ggtitle("8 years")
plot_grid(p4_cord, p4_4years, p4_8years, nrow = 3)
```

This individual is consistently outlier but the difference is very small.


# Residuals

Second, we will compare the results obtained computing the epimutations in the dataset after applying the residuals.


```{r}
rm(list = ls())
## 0 years
load("../INMA_comb.normalizedComBat.autosomic.filterAnnotatedProbes.withNA.PCAresiduals.GenomicRatioSet.Rdata")

## 4 years
load("../INMA4.normalizedComBat.autosomic.filterAnnotatedProbes.withNA.PCAresiduals.GenomicRatioSet.Rdata")


## 8 years
load("../results/epimutations/HELIX.normalizedComBat.autosomic.filterAnnotatedProbes.withNA.PCAresiduals.GenomicRatioSet.Rdata")
```

```{r}
make_res_df <- function(res.list, pheno, dataset){
  nMethod <- sapply(res.list, nrow)
  nMethod[sapply(nMethod, is.null)] <- 0
  methods <- names(res.list)
  names(methods) <- methods
  Reduce(rbind, res.list) %>%
    mutate(method = rep(methods, unlist(nMethod)),
           dataset = dataset) %>%
    left_join(pheno %>% 
                data.frame() %>% 
                select(Sample_Name, Sex, idnum) %>% 
                mutate(sample = Sample_Name,
                       idnum = as.character(idnum)), by = "sample")
}
make_sum_df <- function(res_df){
  res_df %>%
  group_by(method, sample, Sex, idnum, dataset) %>%
  summarize(n = sum(start != 0 & (is.na(pvalue) | pvalue < 0.05/40408))) %>%
  mutate(n_cat = ifelse(n > 1, "1+", as.character(n)))
}

```

```{r}
load("../INMA_comb.epimutations.allSamples.residuals.Rdata")
load("../INMA4.epimutations.allSamples.residuals.Rdata")
load("../results/epimutations/HELIX.epimutations.allSamples.residuals.Rdata")

helixp <- colData(helix)
helixp$idnum <- sapply(strsplit(helixp$SampleID, "_"), `[`, 2)
helixp$Sex <- helixp$e3_sex
helixp$Sample_Name <- helixp$SampleID

res.inma0.df <- make_res_df(res.inma0.residuals.list, pheno = colData(inma0), dataset = "Cord Blood") 
inma0.sum.df <- make_sum_df(res.inma0.df)


res.inma4.df <- make_res_df(res.inma4.residuals.list, pheno = colData(inma4), dataset = "4 years") 
inma4.sum.df <- make_sum_df(res.inma4.df)

res.helix.df <- make_res_df(res.helix.residuals.list, pheno = helixp, dataset = "8 years") 
helix.sum.df <- make_sum_df(res.helix.df)

all.sum.df <- rbind(inma0.sum.df, inma4.sum.df, helix.sum.df) %>%
  mutate(dataset = factor(dataset, levels = c("Cord Blood", "4 years", "8 years")))
methods <- unique(all.sum.df$method)
names(methods ) <- methods

all.res.df <- rbind(res.inma0.df, res.inma4.df, res.helix.df) %>%
  mutate(dataset = factor(dataset, levels = c("Cord Blood", "4 years", "8 years")))
```

```{r}
all.sum.df %>%
  ggplot(aes(x = n_cat)) +
  geom_bar() +
  theme_bw() +
  facet_grid(dataset ~ method, scales = "free_y")

sum.props <- lapply(methods, function(x) {
  
  sub <- subset(all.sum.df, method == x)
   round(prop.table(table(sub$dataset, sub$n_cat), margin = 1)*100, 2)

})
sum.props
```

We observe the same trends than with the raw dataset, where at cord blood more children have more epimutations and more children have at least one epimutation.

```{r}
sab.sum.df <- subset(all.sum.df, dataset != "8 years" | grepl("SAB", sample))

tab <- table(sab.sum.df$idnum)
dup <- names(tab)[tab > 4]
com.res.df <- subset(all.res.df, idnum %in% dup & chromosome != 0 & (is.na(pvalue) | pvalue < 0.05/40408))
long_sum <- com.res.df %>%
  group_by(idnum, method, epi_region_id) %>%
  summarize(n_dataset = length(unique(dataset))) %>%
  filter(n_dataset > 1) %>%
  arrange(method)

table(long_sum$method, long_sum$n_dataset)
```

After applying the residuals, we detected epimutations more constant along time-points. In this case, we also detect epimutaitons shared between 0 and 4 years with barbosa, more epimutations are common across two time-points in beta and even one is common in three time-points.


```{r}
epi_com_beta <- subset(com.res.df, idnum == "304" & method %in% c("beta", "barbosa"))

p1_cord <- plot_epimutations(epi_com_beta[1,], inma0) +
  ggtitle("Cord Blood")
p1_4years <- epi_com_beta %>%
  mutate(sample = "04_304_4") %>%
  plot_epimutations(methy = inma4) +
  ggtitle("4 years")
p1_8years <- epi_com_beta %>%
  mutate(sample = "SAB_304_1X",
         cpg_ids = "cg27182551,cg13431205,cg18481241,cg25676335,cg00059930,cg26740195") %>%
  plot_epimutations(methy = helix) +
  ggtitle("8 years")
plot_grid(p1_cord, p1_4years, p1_8years, nrow = 3)
```

This epimutation was previously detected with beta at cord blood and 4 years in the raw dataset. With the residuals, is also detected by barbosa.



```{r}
epi_com_barbosa <- subset(com.res.df, idnum == "554" & method %in% c("beta", "barbosa") &
                            epi_region_id == "chr16_832683")

p2_cord <- plot_epimutations(epi_com_barbosa[1,], inma0) +
  ggtitle("Cord Blood")
p2_4years <- epi_com_barbosa %>%
  mutate(sample = "04_554_4") %>%
  plot_epimutations(methy = inma4) +
  ggtitle("4 years")
p2_8years <- epi_com_barbosa %>%
  mutate(sample = "SAB_554_1X",
         cpg_ids = "cg11470442,cg03220734,cg08581018,cg02485642,cg08931647") %>%
  plot_epimutations(methy = helix) +
  ggtitle("8 years")
plot_grid(p2_cord, p2_4years, p2_8years, nrow = 3)
```

This epimutation is observed at cord blood and 4 years, while at 8 years the individual have value similar to the population.

```{r}
epi_com_beta2 <- subset(com.res.df, idnum == "294" & method %in% c("beta", "barbosa") &
                         epi_region_id == "chr8_42234243")

p3_cord <- plot_epimutations(epi_com_beta2[1,], inma0) +
  ggtitle("Cord Blood")
p3_4years <- epi_com_beta2 %>%
  mutate(sample = "04_294_4") %>%
  plot_epimutations(methy = inma4) +
  ggtitle("4 years")
p3_8years <- epi_com_beta2 %>%
  mutate(sample = "SAB_294_1X") %>%
  plot_epimutations(methy = helix) +
  ggtitle("8 years")
plot_grid(p3_cord, p3_4years, p3_8years, nrow = 3)
```

This epimutation is clearer at 0 and 4 years than at 8 years.