---
title: "HELIX epimutations residuals"
author: "Carlos Ruiz"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Introduction

In this document, we will analyze the epimutations identified in HELIX We used four methods to detected epimutations (beta, barbosa, mlm and manova). We compared each samples against the rest of the cohort (leave one out). We run the analysis including all samples, only boys or only girls and each cohort analyzed independently. 


```{r}
library(tidyverse)
library(cowplot)
library(minfi)
library(robustbase)
library(epimutacions)

load("../results/preprocess/HELIX/HELIX.withNA.GenomicRatioSet.Rdata")
helix.ori <- gset[, gset$cohort != "MOBA" & gset$h_ethnicity_3cat == "WhiteEur_WhiteOther"]
load("../results/epimutations/HELIX.normalizedComBat.autosomic.filterAnnotatedProbes.withNA.PCAresiduals.GenomicRatioSet.Rdata")

```

```{r}
make_res_df <- function(res.list){
  nMethod <- sapply(res.list, nrow)
  nMethod[sapply(nMethod, is.null)] <- 0
  methods <- names(res.list)
  names(methods) <- methods
  Reduce(rbind, res.list) %>%
    mutate(method = rep(methods, unlist(nMethod))) %>%
    left_join(colData(gset) %>% 
                data.frame() %>% 
                select(SampleID , e3_sex, cohort) %>% 
                mutate(sample = SampleID ), by = "sample")
}
make_sum_df <- function(res_df){
  res_df %>%
  group_by(method, sample, e3_sex, cohort) %>%
  summarize(n = sum(start != 0 & (is.na(pvalue) | pvalue < 0.05/40408))) %>%
  mutate(n_plot = ifelse(n > 10, "10+", as.character(n)),
         n_plot = factor(n_plot, levels = c(0:10, "10+")))
}

```


# All samples

## Summary

We first evaluated the results of each method independently. In all the comparisons, we used a robust poisson model to account for individuals with an extremely large number of epimutations. 

```{r }
load("../results/epimutations/HELIX.epimutations.allSamples.residuals.Rdata")

methods <- names(res.helix.residuals.list)
res.helix.df <- make_res_df(res.helix.residuals.list) 
all.sum.df <- make_sum_df(res.helix.df)

```

```{r}
all.sum.df %>%
  ggplot(aes(x = n_plot, fill = method)) +
  geom_bar(position=position_dodge()) +
  theme_bw()
summary(glmrob(n ~ method, all.sum.df, family = "poisson"))
```

As previously reported, barbosa and beta did not detect epimutations in most of the samples, while mlm and manova detected epimutations in 1/3 of individuals. 

```{r}
all.sum.df %>%
  ggplot(aes(x = n_plot, fill = e3_sex)) +
  geom_bar(position=position_dodge()) +
  theme_bw() +
  facet_wrap(~ method, scales = "free")

lapply(methods, function(x){
  summary(glmrob(n ~ e3_sex, all.sum.df, subset = method == x, family = "poisson"))
})

```

We did not found differences between boys and girls.


```{r}
all.sum.df %>%
  ggplot(aes(x = n_plot, fill = cohort)) +
  geom_bar(position=position_dodge()) +
  theme_bw() +
  facet_wrap(~ method, scales = "free")

lapply(methods, function(x){
  summary(glmrob(n ~ cohort, all.sum.df, subset = method == x, family = "poisson"))
})

```

We did not find differences between the cohorts. 


## Comparison with crude data

```{r }
load("../results/epimutations/HELIX.epimutations.allSamples.Rdata")

res.helix.ori.df <- make_res_df(res.helix.list) 
all.sum.ori.df <- make_sum_df(res.helix.ori.df)

```


```{r}
ori.comb <- all.sum.df %>%
  ungroup() %>%
  mutate(residuals = n) %>%
  select(residuals, method, sample) %>% 
  right_join(all.sum.ori.df %>%
  ungroup() %>%
  mutate(crude = n) %>%
  select(crude, method, sample), by = c("method", "sample")) %>%
  mutate(Epimutations_diff = residuals - crude) 

ggplot(ori.comb, aes(x = Epimutations_diff)) +
  geom_histogram(binwidth = 1) +
  facet_grid(method ~ .) +
  theme_bw() +
  xlab("Difference in epimutations between raw and residuals")
```

With barbosa, we got the same number of epimutations with raw or residuals. For beta, one sample with a high number of epimutations in the raw data, had a reduced number of epimutations, while other samples had slight differences. 

Mlm and manova had a great heterogeinity in the results, with a high proportion of individuals having a different number of epimutations. 

```{r}
getOverlaps <- function(method, sample, epi_df1, epi_df2){
  
  epi1 <- epi_df1[epi_df1$method == method & epi_df1$sample == sample, ]
  epi1 <- subset(epi1, start != 0 & (is.na(pvalue) | pvalue < 0.05/40408))
  
  epi2 <- epi_df2[epi_df2$method == method & epi_df2$sample == sample, ]
  epi2 <- subset(epi2, start != 0 & (is.na(pvalue) | pvalue < 0.05/40408))

  
  n1 <- nrow(epi1)
  n2 <- nrow(epi2)
  
  if (n1 == 0 | n2 == 0){
    return(c(epi1 = n1, epi2 = n2, Common = 0))
  }
  epi1GR <- makeGRangesFromDataFrame(epi1)
  epi2GR <- makeGRangesFromDataFrame(epi2)
  com <- sum(countOverlaps(epi1GR, epi2GR))
  return(c(epi1 = n1 - com, epi2 = n2 - com, Common = com))

}
ori.over <- lapply(seq_len(nrow(ori.comb)), function(i){
  getOverlaps(ori.comb[i, ]$method,
              ori.comb[i, ]$sample,
              res.helix.ori.df, res.helix.df)
})
ori.over.df <- Reduce(rbind, ori.over) %>%
  cbind(ori.comb[, c("method", "sample")], .) %>%
  tibble() %>%
  mutate(Raw = epi1, Residuals = epi2)

ori.over.df %>%
  filter(!(Raw == 0 & Residuals == 0 & Common == 0 )) %>%
  gather(Type, N, 5:7) %>%
  mutate(Type = factor(Type, levels = c("Raw", "Residuals", "Common"))) %>%
  mutate(N = ifelse(N > 25, 25, N)) %>%
  ggplot(aes(x = N, fill = Type)) +
  geom_histogram(binwidth = 1) +
  facet_grid(method ~ .) +
  theme_bw()
```

In this plot, we show the epimutations detected only in the raw dataset, only in the residuals and in common (raw and residuals) for those samples where at least one epimutation was detected. For barbosa, no individuals had epimutations specific of raw or residuals datasets. In beta, we observed a higher proportion of individuals with epimutations specific of the raw dataset. 

Finally, for mlm and manova we observed a high heterogeneity, with many individuals having epimutations specific of the raw or residuals dataset. 

```{r}


epi_crude <- subset(res.helix.df, sample == "KAN_1051_1X" & method %in% c("beta", "barbosa"))
epi_crude2 <- subset(res.helix.ori.df, sample == "KAN_1051_1X" & method %in% c("beta", "barbosa"))
p1_crude <- plot_epimutations(epi_crude2[2,], helix.ori) +
  ggtitle("Raw values")
p1_residuals <- plot_epimutations(epi_crude2[2,], helix) +
  ggtitle("Residual values")
plot_grid(p1_crude, p1_residuals, ncol = 2)

epi_crude <- subset(res.helix.df, sample == "BIB_12153_1X" & method %in% c("beta", "barbosa"))
epi_crude2 <- subset(res.helix.ori.df, sample == "BIB_12153_1X" & method %in% c("beta", "barbosa"))
p2_crude <- plot_epimutations(epi_crude[1,], helix.ori) +
  ggtitle("Raw values")
p2_residuals <- plot_epimutations(epi_crude[1,], helix) +
  ggtitle("Residual values")
plot_grid(p2_crude, p2_residuals, ncol = 2)
```


These two examples show the difference two epimutations only detected in the raw or the residuals. 

# Stratification by sex

We evaluated the effect of running the epimutations in boys and girls independently.

## Boys

```{r }
load("../results/epimutations/HELIX.epimutations.boys.residuals.Rdata")

res.helix.boys.df <- make_res_df(res.helix.boys.residuals.list) 
boys.sum.df <- make_sum_df(res.helix.boys.df)
```


```{r}
boys.comb <- all.sum.df %>%
  ungroup() %>%
  mutate(all = n) %>%
  select(all, method, sample) %>% 
  right_join(boys.sum.df %>%
  ungroup() %>%
  mutate(boys = n) %>%
  select(boys, method, sample), by = c("method", "sample")) %>%
  mutate(Epimutations_diff = boys - all) 

ggplot(boys.comb, aes(x = Epimutations_diff)) +
  geom_histogram(binwidth = 1) +
  facet_grid(method ~ .) +
  theme_bw() +
  xlab("Difference in epimutations between boys and all")
```

For beta and barbosa, only one sample had fewer epimutations when analyizing only boys. For mlm and manova, we have a very variable number of epimutations different. 

```{r}
boys.over <- lapply(seq_len(nrow(boys.comb)), function(i){
  getOverlaps(boys.comb[i, ]$method,
              boys.comb[i, ]$sample,
              res.helix.boys.df, res.helix.df)
})
boys.over.df <- Reduce(rbind, boys.over) %>%
  cbind(boys.comb[, c("method", "sample")], .) %>%
  tibble() %>%
  mutate(Boys = epi1, All = epi2)

boys.over.df %>%
  filter(!(All == 0 & Boys == 0 & Common == 0 )) %>%
  gather(Type, N, 5:7) %>%
  mutate(N = ifelse(N > 25, 25, N)) %>%
  ggplot(aes(x = N, fill = Type)) +
  geom_histogram(binwidth = 1) +
  facet_grid(method ~ .) +
  theme_bw()
```

As in the raw dataset, we have very few differences between analyzing the whole cohort or onyl boys with beta and barbosa, while we have very different epimutations for mlm and manova. 

## Boys in crude model

We explored whether epimutations specific of boys were detected when using the models with residuals.

```{r }
load("../results/epimutations/HELIX.epimutations.boys.Rdata")

res.helix.boys.ori.df <- make_res_df(res.helix.boys.list) 
boys.sum.ori.df <- make_sum_df(res.helix.boys.ori.df)
```

```{r}
getSpecific <- function(method, sample, epi_df1, epi_df2){
  epi1 <- epi_df1[epi_df1$method == method & epi_df1$sample == sample, ]
  epi1 <- subset(epi1, start != 0 & (is.na(pvalue) | pvalue < 0.05/40408))
  
  epi2 <- epi_df2[epi_df2$method == method & epi_df2$sample == sample, ]
  epi2 <- subset(epi2, start != 0 & (is.na(pvalue) | pvalue < 0.05/40408))

  n1 <- nrow(epi1)
  n2 <- nrow(epi2)
  
  if (n1 == 0 | n2 == 0){
    return(epi1)
  }
  epi1GR <- makeGRangesFromDataFrame(epi1)
  epi2GR <- makeGRangesFromDataFrame(epi2)
  ov <- findOverlaps(epi1GR, epi2GR)
  return(epi1[-from(unique(ov)), ])
}
boys.specific.ori <- lapply(seq_len(nrow(boys.comb)), function(i){
  getSpecific(boys.comb[i, ]$method,
              boys.comb[i, ]$sample,
              res.helix.boys.ori.df, res.helix.ori.df)
}) %>% Reduce(f = rbind, x = .)

boys.pairs <- boys.specific.ori %>%
  select(method, sample) %>%
  distinct()

boys.specific.over <- lapply(seq_len(nrow(boys.pairs)), function(i){
  getOverlaps(boys.pairs[i, ]$method,
              boys.pairs[i, ]$sample,
              boys.specific.ori, res.helix.df)
})
Reduce(rbind, boys.specific.over) %>%
  cbind(boys.pairs, .) %>%
  tibble() %>%
  mutate(Boys_ori = epi1,
         Boys_ori_tot = Boys_ori + Common) %>%
  group_by(method) %>%
  summarize(prop = sum(Common)/sum(Boys_ori_tot))

```

This values represent the proportion of epimutations specific of boys in the raw dataset that are captured in the whole dataset after applying the residuals. Due to the small differences of the stratified analyses, we did not find any new epimutation for barbosa or beta. For mlm and manova, we have a low percentage, suggesting that adjusting for residuals do not increase the epimutations specific of boys.

```{r}
lapply(seq_len(nrow(boys.pairs)), function(i){
  getOverlaps(boys.pairs[i, ]$method,
              boys.pairs[i, ]$sample,
              boys.specific.ori, res.helix.boys.df)
}) %>% Reduce(f = rbind, x = .) %>%
  cbind(boys.pairs, .) %>%
  tibble() %>%
  mutate(Boys_ori = epi1,
         Boys_ori_tot = Boys_ori + Common) %>%
  group_by(method) %>%
  summarize(prop = sum(Common)/sum(Boys_ori_tot))

```


Around 20% of epimutations specific of boys in mlm and manova are also specific of boys in the residuals model. These results suggest that epimutations only identified in boys are not due to global different levels in methylation between boys and girls. 

## Girls

```{r }
load("../results/epimutations/HELIX.epimutations.girls.residuals.Rdata")

res.helix.girls.df <- make_res_df(res.helix.girls.residuals.list) 
girls.sum.df <- make_sum_df(res.helix.girls.df)
```


```{r}
girls.comb <- all.sum.df %>%
  ungroup() %>%
  mutate(all = n) %>%
  select(all, method, sample) %>% 
  right_join(girls.sum.df %>%
  ungroup() %>%
  mutate(girls = n) %>%
  select(girls, method, sample), by = c("method", "sample")) %>%
  mutate(Epimutations_diff = girls - all) 

ggplot(girls.comb, aes(x = Epimutations_diff)) +
  geom_histogram(binwidth = 1) +
  facet_grid(method ~ .) +
  theme_bw() +
  xlab("Difference in epimutations between girls and all")
```

Few samples have a different number of epimutations only in girls than in the whole cohort, for beta and barbosa. For mlm and manova, we observe a high difference.

```{r}
girls.over <- lapply(seq_len(nrow(girls.comb)), function(i){
  getOverlaps(girls.comb[i, ]$method,
              girls.comb[i, ]$sample,
              res.helix.girls.df, res.helix.df)
})
girls.over.df <- Reduce(rbind, girls.over) %>%
  cbind(girls.comb[, c("method", "sample")], .) %>%
  tibble() %>%
  mutate(Girls = epi1, All = epi2)

girls.over.df %>%
  filter(!(All == 0 & Girls == 0 & Common == 0 )) %>%
  gather(Type, N, 5:7) %>%
  mutate(Type = factor(Type, levels = c("All", "Girls", "Common"))) %>%
  mutate(N = ifelse(N > 25, 25, N)) %>%
  ggplot(aes(x = N, fill = Type)) +
  geom_histogram(binwidth = 1) +
  facet_grid(method ~ .) +
  theme_bw()
```

Barbosa and beta have so few epimutations that no differnces are seen. mlm seem to have more coherent results between boys and whole dataset when using residuals. 

## Girls in crude model

We explored whether epimutations specific of girls were detected when using the models with residuals.

```{r }
load("../results/epimutations/HELIX.epimutations.girls.Rdata")

res.helix.girls.ori.df <- make_res_df(res.helix.girls.list) 
girls.sum.ori.df <- make_sum_df(res.helix.girls.ori.df)
```

```{r}
girls.specific.ori <- lapply(seq_len(nrow(girls.comb)), function(i){
  getSpecific(girls.comb[i, ]$method,
              girls.comb[i, ]$sample,
              res.helix.girls.ori.df, res.helix.ori.df)
}) %>% Reduce(f = rbind, x = .)

girls.pairs <- girls.specific.ori %>%
  select(method, sample) %>%
  distinct()

girls.specific.over <- lapply(seq_len(nrow(girls.pairs)), function(i){
  getOverlaps(girls.pairs[i, ]$method,
              girls.pairs[i, ]$sample,
              girls.specific.ori, res.helix.df)
})
Reduce(rbind, girls.specific.over) %>%
  cbind(girls.pairs, .) %>%
  tibble() %>%
  mutate(girls_ori = epi1,
         girls_ori_tot = girls_ori + Common) %>%
  group_by(method) %>%
  summarize(prop = sum(Common)/sum(girls_ori_tot))

```

These values represent the proportion of epimutations specific of girls in the raw dataset that are captured in the whole dataset after applying the residuals. For all the methods, this proportion is very low, suggesting that adjusting for residuals do not increase the epimutations specific of girls.

```{r}
lapply(seq_len(nrow(girls.pairs)), function(i){
  getOverlaps(girls.pairs[i, ]$method,
              girls.pairs[i, ]$sample,
              girls.specific.ori, res.helix.girls.df)
}) %>% Reduce(f = rbind, x = .) %>%
  cbind(girls.pairs, .) %>%
  tibble() %>%
  mutate(Girls_ori = epi1,
         Girls_ori_tot = Girls_ori + Common) %>%
  group_by(method) %>%
  summarize(prop = sum(Common)/sum(Girls_ori_tot))

```

For barbosa and beta, none of the epimutations specific of girls in the raw dataset, were also detected only in girls in the residuals dataset. For mlm and manova, these percentages were around 20 %.

These results suggest that epimutations only identified in girls are not due to global different levels in methylation between boys and girls. 


# Stratification by cohort

We evaluated the effect of running the epimutations in each batch independently.

## Esteller

```{r }
load("../results/epimutations/HELIX.epimutations.cohort.residuals.Rdata")

comb <- function(x, y) {
  res <- lapply(names(x), function(z) rbind(x[[z]], y[[z]]))
  names(res) <- names(x)
  res
}
res.helix.cohort.list.comb <- Reduce(comb, res.helix.cohort.residuals.list)
res.helix.cohort.df <- make_res_df(res.helix.cohort.list.comb) 
cohort.sum.df <- make_sum_df(res.helix.cohort.df)
```


```{r}
cohort.comb <- all.sum.df %>%
  ungroup() %>%
  mutate(all = n) %>%
  select(all, method, sample) %>% 
  right_join(cohort.sum.df %>%
  ungroup() %>%
  mutate(cohort = n) %>%
  select(cohort, method, sample), by = c("method", "sample")) %>%
  mutate(Epimutations_diff = cohort - all) 

ggplot(cohort.comb, aes(x = Epimutations_diff)) +
  geom_histogram(binwidth = 1) +
  facet_grid(method ~ .) +
  theme_bw() +
  xlab("Difference in epimutations between cohort and all")
```

After applying the residuals, we still see some individulas with more epimutations in the cohort alone analysis for beta and barbosa. For mlm and manova, we see a high number differences. 


```{r}
cohort.over <- lapply(seq_len(nrow(cohort.comb)), function(i){
  getOverlaps(cohort.comb[i, ]$method,
              cohort.comb[i, ]$sample,
              res.helix.cohort.df, res.helix.df)
})
cohort.over.df <- Reduce(rbind, cohort.over) %>%
  cbind(cohort.comb[, c("method", "sample")], .) %>%
  tibble() %>%
  mutate(Cohort = epi1, All = epi2)

cohort.over.df %>%
  filter(!(All == 0 & Cohort == 0 & Common == 0 )) %>%
  gather(Type, N, 5:7) %>%
  mutate(Type = factor(Type, levels = c("All", "Cohort", "Common"))) %>%
  mutate(N = ifelse(N > 25, 25, N)) %>%
  ggplot(aes(x = N, fill = Type)) +
  geom_histogram(binwidth = 1) +
  facet_grid(method ~ .) +
  theme_bw()
```

When comparing the epimtuations identified in the whole dataset or stratified by cohort, the conclussions are very similar to the raw dataset. Beta and barbosa are very stable, and only few samples have more epimutations in cohort stratified anlaysis. On the other hand, mlm and manova have the more inestable results, and smaples can have either more or less epimutations.

