---
title: "HELIX epimutations"
author: "Carlos Ruiz"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Introduction

In this document, we will analyze the epimutations identified in HELIX We used four methods to detected epimutations (beta, barbosa, mlm and manova). We compared each samples against the rest of the cohort (leave one out). We run the analysis including all samples, only boys or only girls. 


```{r}
library(tidyverse)
library(cowplot)
library(minfi)
library(robustbase)
library(epimutacions)

load("../results/preprocess/HELIX/HELIX.withNA.GenomicRatioSet.Rdata")
helix <- gset[, gset$cohort != "MOBA" & gset$h_ethnicity_3cat == "WhiteEur_WhiteOther"]

```

```{r}
make_res_df <- function(res.list){
  nMethod <- sapply(res.list, nrow)
  nMethod[sapply(nMethod, is.null)] <- 0
  methods <- names(res.list)
  names(methods) <- methods
  Reduce(rbind, res.list) %>%
    mutate(method = rep(methods, unlist(nMethod))) %>%
    left_join(colData(gset) %>% 
                data.frame() %>% 
                select(SampleID , e3_sex, cohort) %>% 
                mutate(sample = SampleID ), by = "sample")
}
make_sum_df <- function(res_df){
  res_df %>%
  group_by(method, sample, e3_sex, cohort) %>%
  summarize(n = sum(start != 0 & (is.na(pvalue) | pvalue < 0.05/40408))) %>%
  mutate(n_plot = ifelse(n > 10, "10+", as.character(n)),
         n_plot = factor(n_plot, levels = c(0:10, "10+")))
}

```

# All samples

## Individual methods

We first evaluated the results of each method independently. In all the comparisons, we used a robust poisson model to account for individuals with an extremely large number of epimutations. 

```{r }
load("../results/epimutations/HELIX.epimutations.allSamples.Rdata")

methods <- names(res.helix.list)
res.helix.df <- make_res_df(res.helix.list) 
all.sum.df <- make_sum_df(res.helix.df)

```

```{r}
all.sum.df %>%
  ggplot(aes(x = n_plot, fill = method)) +
  geom_bar(position=position_dodge()) +
  theme_bw()
summary(glmrob(n ~ method, all.sum.df, family = "poisson"))
```

As previously reported, barbosa is the method that detects less epimutations, and only few samples had an epimutation. Beta detected epimutations in more individuals while around 1/3 of individuals had epimutations with mlm or manova.

```{r}
all.sum.df %>%
  ggplot(aes(x = n_plot, fill = e3_sex)) +
  geom_bar(position=position_dodge()) +
  theme_bw() +
  facet_wrap(~ method, scales = "free")

lapply(methods, function(x){
  summary(glmrob(n ~ e3_sex, all.sum.df, subset = method == x, family = "poisson"))
})

```

We did not found differences between boys and girls.


```{r}
all.sum.df %>%
  ggplot(aes(x = n_plot, fill = cohort)) +
  geom_bar(position=position_dodge()) +
  theme_bw() +
  facet_wrap(~ method, scales = "free")

lapply(methods[-2], function(x){
  summary(glmrob(n ~ cohort, all.sum.df, subset = method == x, family = "poisson"))
})

```

We did not find differences in the number of epimutations by cohort. 

# Stratification by sex

We evaluated the effect of running the epimutations in boys and girls independently.

## Boys

```{r }
load("../results/epimutations/HELIX.epimutations.boys.Rdata")

res.helix.boys.df <- make_res_df(res.helix.boys.list) 
boys.sum.df <- make_sum_df(res.helix.boys.df)
```


```{r}
boys.comb <- all.sum.df %>%
  ungroup() %>%
  mutate(all = n) %>%
  select(all, method, sample) %>% 
  right_join(boys.sum.df %>%
  ungroup() %>%
  mutate(boys = n) %>%
  select(boys, method, sample), by = c("method", "sample")) %>%
  mutate(Epimutations_diff = boys - all) 

ggplot(boys.comb, aes(x = Epimutations_diff)) +
  geom_histogram(binwidth = 1) +
  facet_grid(method ~ .) +
  theme_bw() +
  xlab("Difference in epimutations between boys and all")
```

We observed exactly the same number of epimutations for beta and barbosa when analyzing the whole cohort or when only analyzing boys.  For mlm and manova, some individuals have few more epimutations in one of the analyses.

```{r}
getOverlaps <- function(method, sample, epi_df1, epi_df2){
  
  epi1 <- epi_df1[epi_df1$method == method & epi_df1$sample == sample, ]
  epi1 <- subset(epi1, start != 0 & (is.na(pvalue) | pvalue < 0.05/40408))
  
  epi2 <- epi_df2[epi_df2$method == method & epi_df2$sample == sample, ]
  epi2 <- subset(epi2, start != 0 & (is.na(pvalue) | pvalue < 0.05/40408))

  
  n1 <- nrow(epi1)
  n2 <- nrow(epi2)
  
  if (n1 == 0 | n2 == 0){
    return(c(epi1 = n1, epi2 = n2, Common = 0))
  }
  epi1GR <- makeGRangesFromDataFrame(epi1)
  epi2GR <- makeGRangesFromDataFrame(epi2)
  com <- sum(countOverlaps(epi1GR, epi2GR))
  return(c(epi1 = n1 - com, epi2 = n2 - com, Common = com))

}
boys.over <- lapply(seq_len(nrow(boys.comb)), function(i){
  getOverlaps(boys.comb[i, ]$method,
              boys.comb[i, ]$sample,
              res.helix.boys.df, res.helix.df)
})
boys.over.df <- Reduce(rbind, boys.over) %>%
  cbind(boys.comb[, c("method", "sample")], .) %>%
  tibble() %>%
  mutate(Boys = epi1, All = epi2)

boys.over.df %>%
  filter(!(All == 0 & Boys == 0 & Common == 0 )) %>%
  gather(Type, N, 5:7) %>%
  mutate(N = ifelse(N > 25, 25, N)) %>%
  ggplot(aes(x = N, fill = Type)) +
  geom_histogram(binwidth = 1) +
  facet_grid(method ~ .) +
  theme_bw()
```

In this plot, we show the epimutations detected only in boys, only in all and in common (boys and all) for those samples where at least one epimutation was detected. As barbosa and beta detected epimutations in very few individuals, the columns are very small. For these methods, the epimutations detected in the whole dataset are also detected in boys. 

We observed a high heterogeneity for mlm and manova  with many individuals having epimutations specific of the boys or all datasets.

## Girls


```{r }
load("../results/epimutations/HELIX.epimutations.girls.Rdata")

res.helix.girls.df <- make_res_df(res.helix.girls.list) 
girls.sum.df <- make_sum_df(res.helix.girls.df)
```



```{r}
girls.comb <- all.sum.df %>%
  ungroup() %>%
  mutate(all = n) %>%
  select(all, method, sample) %>% 
  right_join(girls.sum.df %>%
  ungroup() %>%
  mutate(girls = n) %>%
  select(girls, method, sample), by = c("method", "sample")) %>%
  mutate(Epimutations_diff = girls - all) 

ggplot(girls.comb, aes(x = Epimutations_diff)) +
  geom_histogram(binwidth = 1) +
  facet_grid(method ~ .) +
  theme_bw() +
  xlab("Difference in epimutations between girls and all")
```

The general are the same than for boys. The only difference is a girl that had two epimutations more when analyzing girls alone.

```{r}
girls.over <- lapply(seq_len(nrow(girls.comb)), function(i){
  getOverlaps(girls.comb[i, ]$method,
              girls.comb[i, ]$sample,
              res.helix.girls.df, res.helix.df)
})
girls.over.df <- Reduce(rbind, girls.over) %>%
  cbind(girls.comb[, c("method", "sample")], .) %>%
  tibble() %>%
  mutate(Girls = epi1, All = epi2)

girls.over.df %>%
  filter(!(All == 0 & Girls == 0 & Common == 0 )) %>%
  gather(Type, N, 5:7) %>%
  mutate(Type = factor(Type, levels = c("All", "Girls", "Common"))) %>%
  mutate(N = ifelse(N > 25, 25, N)) %>%
  ggplot(aes(x = N, fill = Type)) +
  geom_histogram(binwidth = 1) +
  facet_grid(method ~ .) +
  theme_bw()
```

In this plot, we show the epimutations detected only in girls, only in all and in common (girls and all) for those samples where at least one epimutation was detected. The trends are very similar than for boys: barbosa presented no differences between analyzing the whole dataset or the girls alone, beta presented few epimutations specific of girls or of all dataset, while mlm and manova presented high heterogeneity, with many individuals having epimutations specific of the girls or all datasets.

# Stratification by cohort

We evaluated whether the different cohorts had a different number of epimutations.


```{r}
load("../results/epimutations/HELIX.epimutations.cohort.Rdata")

comb <- function(x, y) {
  res <- lapply(names(x), function(z) rbind(x[[z]], y[[z]]))
  names(res) <- names(x)
  res
}
res.helix.cohort.list.comb <- Reduce(comb, res.helix.cohort.list)
res.helix.cohort.df <- make_res_df(res.helix.cohort.list.comb) 
cohort.sum.df <- make_sum_df(res.helix.cohort.df)
```


```{r}
cohort.comb <- all.sum.df %>%
  ungroup() %>%
  mutate(all = n) %>%
  select(all, method, sample) %>% 
  right_join(cohort.sum.df %>%
  ungroup() %>%
  mutate(cohort = n) %>%
  select(cohort, method, sample), by = c("method", "sample")) %>%
  mutate(Epimutations_diff = cohort - all) 

ggplot(cohort.comb, aes(x = Epimutations_diff)) +
  geom_histogram(binwidth = 1) +
  facet_grid(method ~ .) +
  theme_bw() +
  xlab("Difference in epimutations between cohort and all")
```

For beta and barbosa, most individuals had the same number of epimutations when runing the whole cohort or when running the analysis by cohort. Again, we see more discrepancies for mlm and manova.


```{r}
cohort.over <- lapply(seq_len(nrow(cohort.comb)), function(i){
  getOverlaps(cohort.comb[i, ]$method,
              cohort.comb[i, ]$sample,
              res.helix.cohort.df, res.helix.df)
})
cohort.over.df <- Reduce(rbind, cohort.over) %>%
  cbind(cohort.comb[, c("method", "sample")], .) %>%
  tibble() %>%
  mutate(Cohort = epi1, All = epi2)

cohort.over.df %>%
  filter(!(All == 0 & Cohort == 0 & Common == 0 )) %>%
  gather(Type, N, 5:7) %>%
  mutate(Type = factor(Type, levels = c("All", "Cohort", "Common"))) %>%
  mutate(N = ifelse(N > 25, 25, N)) %>%
  ggplot(aes(x = N, fill = Type)) +
  geom_histogram(binwidth = 1) +
  facet_grid(method ~ .) +
  theme_bw()
```

For barbosa and beta, some individuals have more epimutations in the cohort alone analysis. mlm and manova have a great difference of epimutations in the stratified or whole cohort analysis. 


```{r}
epi_cohort <- subset(res.helix.cohort.df, sample == "BIB_12641_1X" & method %in% c("beta", "barbosa"))

p2_bib <- plot_epimutations(epi_cohort[2,], helix[, helix$cohort == "BIB"]) +
  ggtitle("Only BIB")
p2_all <- plot_epimutations(epi_cohort[2,], helix) +
  ggtitle("Whole dataset")
plot_grid(p2_bib, p2_all, ncol = 2)
```

In this figure, we show an epimutation that it is only detected in BIB. This individual is an outlier when considering only BIB butin the whole cohort, more individuals have similar methylation values. 