---
title: "INMA epimutations residuals"
author: "Carlos Ruiz"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Introduction

In this document, we will analyze the epimutations identified in INMA. We used four methods to detected epimutations (beta, barbosa, mlm and manova). We compared each samples against the rest of the cohort (leave one out). We run the analysis including all samples, only boys or only girls. 


```{r}
library(tidyverse)
library(cowplot)
library(minfi)
library(robustbase)
library(epimutacions)

load("../MeDALL_all.normalizedComBat.autosomic.filterAnnotatedProbes.withNA.GenomicRatioSet.Rdata")

gset.4 <- gset[, gset$Sample_Group == "Age_4"]
              
## Select INMA SAB samples
gset.sab <- gset.4[, grep("^04|SAB", gset.4$Sample_Name)]

## Remove duplicates
gset <- gset.sab[, !duplicated(gset.sab$idnum)]
load("../INMA4.normalizedComBat.autosomic.filterAnnotatedProbes.withNA.PCAresiduals.GenomicRatioSet.Rdata")

```

```{r}
make_res_df <- function(res.list){
  nMethod <- sapply(res.list, nrow)
  nMethod[sapply(nMethod, is.null)] <- 0
  methods <- names(res.list)
  names(methods) <- methods
  Reduce(rbind, res.list) %>%
    mutate(method = rep(methods, unlist(nMethod))) %>%
    left_join(colData(gset) %>% 
                data.frame() %>% 
                select(Sample_Name, Sex, smkpreg) %>% 
                mutate(sample = Sample_Name), by = "sample")
}
make_sum_df <- function(res_df){
  res_df %>%
  group_by(method, sample, Sex, smkpreg) %>%
  summarize(n = sum(start != 0 & (is.na(pvalue) | pvalue < 0.05/40408))) %>%
  mutate(n_plot = ifelse(n > 10, "10+", as.character(n)),
         n_plot = factor(n_plot, levels = c(0:10, "10+")))
}

```


# All samples

## Summary

We first evaluated the results of each method independently. In all the comparisons, we used a robust poisson model to account for individuals with an extremely large number of epimutations. 

```{r }
load("../INMA4.epimutations.allSamples.residuals.Rdata")

methods <- names(res.inma4.residuals.list)

res.inma4.df <- make_res_df(res.inma4.residuals.list) 
all.sum.df <- make_sum_df(res.inma4.df)

```

```{r}
all.sum.df %>%
  ggplot(aes(x = n_plot, fill = method)) +
  geom_bar(position=position_dodge()) +
  theme_bw()
summary(glmrob(n ~ method, all.sum.df, family = "poisson"))
```

As previously reported, barbosa is the method that detects less epimutations, followed by beta, mlm and manova. 

```{r}
all.sum.df %>%
  ggplot(aes(x = n_plot, fill = Sex)) +
  geom_bar(position=position_dodge()) +
  theme_bw() +
  facet_wrap(~ method, scales = "free")

lapply(methods, function(x){
  summary(glmrob(n ~ Sex, all.sum.df, subset = method == x, family = "poisson"))
})

```

We did not found differences between boys and girls.


```{r}
all.sum.df %>%
  filter(!is.na(smkpreg)) %>%
  ggplot(aes(x = n_plot, fill = smkpreg)) +
  geom_bar(position=position_dodge()) +
  theme_bw() +
  facet_wrap(~ method, scales = "free")

lapply(methods, function(x){
  summary(glmrob(n ~ smkpreg, all.sum.df, subset = method == x, family = "poisson"))
})
lapply(methods, function(x){
  summary(glmrob(n ~ smkpreg + Sex, all.sum.df, subset = method == x, family = "poisson"))
})

```

We did not find more epimutations in children whose mother smoked during pregnancy.

## Comparison with crude data

```{r }
load("../INMA4.epimutations.allSamples.Rdata")

res.inma4.ori.df <- make_res_df(res.inma4.list) 
all.sum.ori.df <- make_sum_df(res.inma4.ori.df)
```


```{r}
ori.comb <- all.sum.df %>%
  ungroup() %>%
  mutate(residuals = n) %>%
  select(residuals, method, sample) %>% 
  right_join(all.sum.ori.df %>%
  ungroup() %>%
  mutate(crude = n) %>%
  select(crude, method, sample), by = c("method", "sample")) %>%
  mutate(Epimutations_diff = residuals - crude) 

ggplot(ori.comb, aes(x = Epimutations_diff)) +
  geom_histogram(binwidth = 1) +
  facet_grid(method ~ .) +
  theme_bw() +
  xlab("Difference in epimutations between raw and residuals")
```

For beta and barbosa, we detected almost the same number of epimutations with raw values or after computing residuals. For beta, some samples that had a huge number of epimutations in the raw data, had much less epimutations in the residuals. For barbosa, we saw the opposite behaviour.

Mlm and manova had a great heterogeinity in the results, with almost all individuals having a different number of epimutations. 

```{r}
getOverlaps <- function(method, sample, epi_df1, epi_df2){
  
  epi1 <- epi_df1[epi_df1$method == method & epi_df1$sample == sample, ]
  epi1 <- subset(epi1, start != 0 & (is.na(pvalue) | pvalue < 0.05/40408))
  
  epi2 <- epi_df2[epi_df2$method == method & epi_df2$sample == sample, ]
  epi2 <- subset(epi2, start != 0 & (is.na(pvalue) | pvalue < 0.05/40408))

  
  n1 <- nrow(epi1)
  n2 <- nrow(epi2)
  
  if (n1 == 0 | n2 == 0){
    return(c(epi1 = n1, epi2 = n2, Common = 0))
  }
  epi1GR <- makeGRangesFromDataFrame(epi1)
  epi2GR <- makeGRangesFromDataFrame(epi2)
  com <- sum(countOverlaps(epi1GR, epi2GR))
  return(c(epi1 = n1 - com, epi2 = n2 - com, Common = com))

}
ori.over <- lapply(seq_len(nrow(ori.comb)), function(i){
  getOverlaps(ori.comb[i, ]$method,
              ori.comb[i, ]$sample,
              res.inma4.ori.df, res.inma4.df)
})
ori.over.df <- Reduce(rbind, ori.over) %>%
  cbind(ori.comb[, c("method", "sample")], .) %>%
  tibble() %>%
  mutate(Raw = epi1, Residuals = epi2)

ori.over.df %>%
  filter(!(Raw == 0 & Residuals == 0 & Common == 0 )) %>%
  gather(Type, N, 5:7) %>%
  mutate(Type = factor(Type, levels = c("Raw", "Residuals", "Common"))) %>%
  mutate(N = ifelse(N > 25, 25, N)) %>%
  ggplot(aes(x = N, fill = Type)) +
  geom_histogram(binwidth = 1) +
  facet_grid(method ~ .) +
  theme_bw()
```

In this plot, we show the epimutations detected only in the raw dataset, only in the residuals and in common (raw and residuals) for those samples where at least one epimutation was detected. For barbosa and beta, some samples without epimutations in the raw dataset, had epimutations in the residuals dataset. 

For mlm and manova we observed a high heterogeneity, but in general the number of epimutations increased in the residuals dataset.

```{r}
epi_crude <- subset(res.inma4.df, sample == "04_362_4" & method %in% c("beta", "barbosa"))
epi_crude2 <- subset(res.inma4.ori.df, sample == "04_362_4" & method %in% c("beta", "barbosa"))

p1_crude <- plot_epimutations(epi_crude[1,], gset) +
  ggtitle("Raw values")
p1_residuals <- plot_epimutations(epi_crude[1,], inma4) +
  ggtitle("Residual values")
plot_grid(p1_crude, p1_residuals, ncol = 2)

epi_crude <- subset(res.inma4.df, sample == "04_005_4" & method %in% c("beta", "barbosa"))
epi_crude2 <- subset(res.inma4.ori.df, sample == "04_005_4" & method %in% c("beta", "barbosa"))
p2_crude <- plot_epimutations(epi_crude2[4,], gset) +
  ggtitle("Raw values")
p2_residuals <- plot_epimutations(epi_crude2[4,], inma4) +
  ggtitle("Residual values")
plot_grid(p1_crude, p1_residuals, ncol = 2)
```


These two examples show the difference two epimutations only detected in the residuals or raw dataset. 




# Stratification by sex

We evaluated the effect of running the epimutations in boys and girls independently.

## Boys

```{r }
load("../INMA4.epimutations.boys.residuals.Rdata")

res.inma4.boys.df <- make_res_df(res.inma4.boys.residuals.list) 
boys.sum.df <- make_sum_df(res.inma4.boys.df)
```


```{r}
boys.comb <- all.sum.df %>%
  ungroup() %>%
  mutate(all = n) %>%
  select(all, method, sample) %>% 
  right_join(boys.sum.df %>%
  ungroup() %>%
  mutate(boys = n) %>%
  select(boys, method, sample), by = c("method", "sample")) %>%
  mutate(Epimutations_diff = boys - all) 

ggplot(boys.comb, aes(x = Epimutations_diff)) +
  geom_histogram(binwidth = 1) +
  facet_grid(method ~ .) +
  theme_bw() +
  xlab("Difference in epimutations between boys and all")
```

We still see more epimutations in boys compared to the whole cohort when using the dataset with residuals. More samples have a different number of epimutations when comparing boys vs the whole cohort than with the raw data.

```{r}
boys.over <- lapply(seq_len(nrow(boys.comb)), function(i){
  getOverlaps(boys.comb[i, ]$method,
              boys.comb[i, ]$sample,
              res.inma4.boys.df, res.inma4.df)
})
boys.over.df <- Reduce(rbind, boys.over) %>%
  cbind(boys.comb[, c("method", "sample")], .) %>%
  tibble() %>%
  mutate(Boys = epi1, All = epi2)

boys.over.df %>%
  filter(!(All == 0 & Boys == 0 & Common == 0 )) %>%
  gather(Type, N, 5:7) %>%
  mutate(N = ifelse(N > 25, 25, N)) %>%
  ggplot(aes(x = N, fill = Type)) +
  geom_histogram(binwidth = 1) +
  facet_grid(method ~ .) +
  theme_bw()
```

In barbosa, we are detecting more epimutations specific of boys when applying the residuals. With beta, we also see more epimutations specific of boys. Finally, manova and mlm seem to have more coherent results between boys and whole datset when using residuals. 

```{r}
epi_boys <- subset(res.inma4.boys.df, sample == "04_318_4" & method %in% c("beta", "barbosa"))

p1_boys <- plot_epimutations(epi_boys[2,], gset[, gset$Sex == "M"]) +
  ggtitle("Only boys")
p1_all <- plot_epimutations(epi_boys[2,], gset) +
  ggtitle("Whole dataset")
plot_grid(p1_boys, p1_all, ncol = 2)

```

We only detect the epimutation in boys because a high number of girls had more outlier values. 

## Boys in crude model

We explored whether epimutations specific of boys were detected when using the models with residuals.

```{r }
load("../INMA4.epimutations.boys.Rdata")

res.inma4.boys.ori.df <- make_res_df(res.inma4.boys.list) 
boys.sum.ori.df <- make_sum_df(res.inma4.boys.ori.df)
```

```{r}
getSpecific <- function(method, sample, epi_df1, epi_df2){
  epi1 <- epi_df1[epi_df1$method == method & epi_df1$sample == sample, ]
  epi1 <- subset(epi1, start != 0 & (is.na(pvalue) | pvalue < 0.05/40408))
  
  epi2 <- epi_df2[epi_df2$method == method & epi_df2$sample == sample, ]
  epi2 <- subset(epi2, start != 0 & (is.na(pvalue) | pvalue < 0.05/40408))

  n1 <- nrow(epi1)
  n2 <- nrow(epi2)
  
  if (n1 == 0 | n2 == 0){
    return(epi1)
  }
  epi1GR <- makeGRangesFromDataFrame(epi1)
  epi2GR <- makeGRangesFromDataFrame(epi2)
  ov <- findOverlaps(epi1GR, epi2GR)
  return(epi1[-from(unique(ov)), ])
}
boys.specific.ori <- lapply(seq_len(nrow(boys.comb)), function(i){
  getSpecific(boys.comb[i, ]$method,
              boys.comb[i, ]$sample,
              res.inma4.boys.ori.df, res.inma4.ori.df)
}) %>% Reduce(f = rbind, x = .)

boys.pairs <- boys.specific.ori %>%
  select(method, sample) %>%
  distinct()

boys.specific.over <- lapply(seq_len(nrow(boys.pairs)), function(i){
  getOverlaps(boys.pairs[i, ]$method,
              boys.pairs[i, ]$sample,
              boys.specific.ori, res.inma4.df)
})
Reduce(rbind, boys.specific.over) %>%
  cbind(boys.pairs, .) %>%
  tibble() %>%
  mutate(Boys_ori = epi1,
         Boys_ori_tot = Boys_ori + Common) %>%
  group_by(method) %>%
  summarize(prop = sum(Common)/sum(Boys_ori_tot))

```

These values represent the proportion of epimutations specific of boys in the raw dataset that are captured in the whole dataset after applying the residuals. For all the methods, this proportion is very low, suggesting that adjusting for residuals do not increase the epimutations specific of boys.

```{r}
lapply(seq_len(nrow(boys.pairs)), function(i){
  getOverlaps(boys.pairs[i, ]$method,
              boys.pairs[i, ]$sample,
              boys.specific.ori, res.inma4.boys.df)
}) %>% Reduce(f = rbind, x = .) %>%
  cbind(boys.pairs, .) %>%
  tibble() %>%
  mutate(Boys_ori = epi1,
         Boys_ori_tot = Boys_ori + Common) %>%
  group_by(method) %>%
  summarize(prop = sum(Common)/sum(Boys_ori_tot))

```

For barbosa and beta, around 40% of epimutations specific of boys in the raw dataset, were also detected only in boys in the residuals dataset. This proportion was lower for mlm and manova. 

These results suggest that epimutations only identified in boys are not due to global different levels in methylation between boys and girls. 


```{r}
epi_boys <- subset(res.inma4.boys.ori.df, sample == "04_318_4" & method %in% c("beta", "barbosa"))

p1_boys <- plot_epimutations(epi_boys[1,], gset[, gset$Sex == "M"]) +
  ggtitle("Boys - Raw")
p1_all <- plot_epimutations(epi_boys[1,], gset) +
  ggtitle("All - Raw")
p1_boys2 <- plot_epimutations(epi_boys[1,], inma4[, inma4$Sex == "M"]) +
  ggtitle("Boys - Residuals")
p1_all2 <- plot_epimutations(epi_boys[1,], inma4) +
  ggtitle("All - Residuals")
plot_grid(p1_boys, p1_all,p1_boys2, p1_all2,ncol = 2)
```


This epimutation was indetified only in boys, in raw data or with residuals.

## Girls

```{r }
load("../INMA4.epimutations.girls.residuals.Rdata")

res.inma4.girls.df <- make_res_df(res.inma4.girls.residuals.list) 
girls.sum.df <- make_sum_df(res.inma4.girls.df)
```


```{r}
girls.comb <- all.sum.df %>%
  ungroup() %>%
  mutate(all = n) %>%
  select(all, method, sample) %>% 
  right_join(girls.sum.df %>%
  ungroup() %>%
  mutate(girls = n) %>%
  select(girls, method, sample), by = c("method", "sample")) %>%
  mutate(Epimutations_diff = girls - all) 

ggplot(girls.comb, aes(x = Epimutations_diff)) +
  geom_histogram(binwidth = 1) +
  facet_grid(method ~ .) +
  theme_bw() +
  xlab("Difference in epimutations between girls and all")
```

We still see more epimutations in girls compared to the whole cohort when using the dataset with residuals, with more discrepancies for mlm or manova. 

```{r}
girls.over <- lapply(seq_len(nrow(girls.comb)), function(i){
  getOverlaps(girls.comb[i, ]$method,
              girls.comb[i, ]$sample,
              res.inma4.girls.df, res.inma4.df)
})
girls.over.df <- Reduce(rbind, girls.over) %>%
  cbind(girls.comb[, c("method", "sample")], .) %>%
  tibble() %>%
  mutate(Girls = epi1, All = epi2)

girls.over.df %>%
  filter(!(All == 0 & Girls == 0 & Common == 0 )) %>%
  gather(Type, N, 5:7) %>%
  mutate(Type = factor(Type, levels = c("All", "Girls", "Common"))) %>%
  mutate(N = ifelse(N > 25, 25, N)) %>%
  ggplot(aes(x = N, fill = Type)) +
  geom_histogram(binwidth = 1) +
  facet_grid(method ~ .) +
  theme_bw()
```

In barbosa and beta, we are detecting more epimutations specific of girls when applying the residuals. Finally, manova and mlm seem to have more coherent results between girls and whole dataset when using residuals. 

```{r}
epi_girls <- subset(res.inma4.girls.df, sample == "04_407_4" & method %in% c("beta", "barbosa"))

p1_girls <- plot_epimutations(epi_girls[1,], inma4[, inma4$Sex == "F"]) +
  ggtitle("Only girls")
p1_all <- plot_epimutations(epi_girls[1,], inma4) +
  ggtitle("Whole dataset")
plot_grid(p1_girls, p1_all, ncol = 2)
```

We only detect the epimutation in girls because a high number of boys had more outlier values. 

## Girls in crude model

We explored whether epimutations specific of girls were detected when using the models with residuals.

```{r }
load("../INMA4.epimutations.girls.Rdata")

res.inma4.girls.ori.df <- make_res_df(res.inma4.girls.list) 
girls.sum.ori.df <- make_sum_df(res.inma4.girls.ori.df)
```

```{r}
girls.specific.ori <- lapply(seq_len(nrow(girls.comb)), function(i){
  getSpecific(girls.comb[i, ]$method,
              girls.comb[i, ]$sample,
              res.inma4.girls.ori.df, res.inma4.ori.df)
}) %>% Reduce(f = rbind, x = .)

girls.pairs <- girls.specific.ori %>%
  select(method, sample) %>%
  distinct()

girls.specific.over <- lapply(seq_len(nrow(girls.pairs)), function(i){
  getOverlaps(girls.pairs[i, ]$method,
              girls.pairs[i, ]$sample,
              girls.specific.ori, res.inma4.df)
})
Reduce(rbind, girls.specific.over) %>%
  cbind(girls.pairs, .) %>%
  tibble() %>%
  mutate(girls_ori = epi1,
         girls_ori_tot = girls_ori + Common) %>%
  group_by(method) %>%
  summarize(prop = sum(Common)/sum(girls_ori_tot))

```

These values represent the proportion of epimutations specific of girls in the raw dataset that are captured in the whole dataset after applying the residuals. For all the methods, this proportion is very low, suggesting that adjusting for residuals do not increase the epimutations specific of girls.

```{r}
lapply(seq_len(nrow(girls.pairs)), function(i){
  getOverlaps(girls.pairs[i, ]$method,
              girls.pairs[i, ]$sample,
              girls.specific.ori, res.inma4.girls.df)
}) %>% Reduce(f = rbind, x = .) %>%
  cbind(girls.pairs, .) %>%
  tibble() %>%
  mutate(Girls_ori = epi1,
         Girls_ori_tot = Girls_ori + Common) %>%
  group_by(method) %>%
  summarize(prop = sum(Common)/sum(Girls_ori_tot))

```

The proportion of epimutations that were specific of girls in raw and residuals datasets is very low too. 
