---
title: "INMA 4 yeras epimutations"
author: "Carlos Ruiz"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Introduction

In this document, we will analyze the epimutations identified in INMA at 4 years. We used four methods to detected epimutations (beta, barbosa, mlm and manova). We compared each samples against the rest of the cohort (leave one out). We run the analysis including all samples, only boys or only girls. 


```{r}
library(tidyverse)
library(cowplot)
library(minfi)
library(robustbase)
library(epimutacions)

load("../MeDALL_all.normalizedComBat.autosomic.filterAnnotatedProbes.withNA.GenomicRatioSet.Rdata")

gset.4 <- gset[, gset$Sample_Group == "Age_4"]
              
## Select INMA SAB samples
gset.sab <- gset.4[, grep("^04|SAB", gset.4$Sample_Name)]

## Remove duplicates
gset <- gset.sab[, !duplicated(gset.sab$idnum)]

```
```{r}
make_res_df <- function(res.list){
  nMethod <- sapply(res.list, nrow)
  nMethod[sapply(nMethod, is.null)] <- 0
  methods <- names(res.list)
  names(methods) <- methods
  Reduce(rbind, res.list) %>%
    mutate(method = rep(methods, unlist(nMethod))) %>%
    left_join(colData(gset) %>% 
                data.frame() %>% 
                select(Sample_Name, Sex, smkpreg) %>% 
                mutate(sample = Sample_Name), by = "sample")
}
make_sum_df <- function(res_df){
  res_df %>%
  group_by(method, sample, Sex, smkpreg) %>%
  summarize(n = sum(start != 0 & (is.na(pvalue) | pvalue < 0.05/40408))) %>%
  mutate(n_plot = ifelse(n > 10, "10+", as.character(n)),
         n_plot = factor(n_plot, levels = c(0:10, "10+")))
}

```

# All samples

## Individual methods

We first evaluated the results of each method independently. In all the comparisons, we used a robust poisson model to account for individuals with an extremely large number of epimutations. 

```{r }
load("../INMA4.epimutations.allSamples.Rdata")

methods <- names(res.inma4.list)
res.inma4.df <- make_res_df(res.inma4.list) 
all.sum.df <- make_sum_df(res.inma4.df)

```

```{r}
all.sum.df %>%
  ggplot(aes(x = n_plot, fill = method)) +
  geom_bar(position=position_dodge()) +
  theme_bw()
summary(glmrob(n ~ method, all.sum.df, family = "poisson"))
```

As previously reported, barbosa is the method that detects less epimutations, followed by beta, mlm and manova. 

```{r}
all.sum.df %>%
  ggplot(aes(x = n_plot, fill = Sex)) +
  geom_bar(position=position_dodge()) +
  theme_bw() +
  facet_wrap(~ method, scales = "free")

lapply(methods, function(x){
  summary(glmrob(n ~ Sex, all.sum.df, subset = method == x, family = "poisson"))
})

```

We did not found differences between boys and girls.


```{r}
all.sum.df %>%
  filter(!is.na(smkpreg)) %>%
  ggplot(aes(x = n_plot, fill = smkpreg)) +
  geom_bar(position=position_dodge()) +
  theme_bw() +
  facet_wrap(~ method, scales = "free")

lapply(methods[-2], function(x){
  summary(glmrob(n ~ smkpreg, all.sum.df, subset = method == x, family = "poisson"))
})
lapply(methods[-2], function(x){
  summary(glmrob(n ~ smkpreg + Sex, all.sum.df, subset = method == x, family = "poisson"))
})

```

We only found and association between maternal smoking and epimutations for the epimutacions detected with manova.


# Stratification by sex

We evaluated the effect of running the epimutations in boys and girls independently.

## Boys

```{r }
load("../INMA4.epimutations.boys.Rdata")

res.inma4.boys.df <- make_res_df(res.inma4.boys.list) 
boys.sum.df <- make_sum_df(res.inma4.boys.df)

```


```{r}
boys.comb <- all.sum.df %>%
  ungroup() %>%
  mutate(all = n) %>%
  select(all, method, sample) %>% 
  right_join(boys.sum.df %>%
  ungroup() %>%
  mutate(boys = n) %>%
  select(boys, method, sample), by = c("method", "sample")) %>%
  mutate(Epimutations_diff = boys - all) 

ggplot(boys.comb, aes(x = Epimutations_diff)) +
  geom_histogram(binwidth = 1) +
  facet_grid(method ~ .) +
  theme_bw() +
  xlab("Difference in epimutations between boys and all")
```

We observed very coincident number of epimutations between the whole cohort or using only boys when applying barbosa or beta. With barbosa, some indiviauls had more epimutations when analyzing boys alone. With beta, few individuals could have more ore less epimutations when analyzing boys alone. For mlm and manova, a high proportion of individuals had a different number of epimutations when considering the whole cohort or just boys. 

```{r}
getOverlaps <- function(method, sample, epi_df1, epi_df2){
  
  epi1 <- epi_df1[epi_df1$method == method & epi_df1$sample == sample, ]
  epi1 <- subset(epi1, start != 0 & (is.na(pvalue) | pvalue < 0.05/40408))
  
  epi2 <- epi_df2[epi_df2$method == method & epi_df2$sample == sample, ]
  epi2 <- subset(epi2, start != 0 & (is.na(pvalue) | pvalue < 0.05/40408))

  
  n1 <- nrow(epi1)
  n2 <- nrow(epi2)
  
  if (n1 == 0 | n2 == 0){
    return(c(epi1 = n1, epi2 = n2, Common = 0))
  }
  epi1GR <- makeGRangesFromDataFrame(epi1)
  epi2GR <- makeGRangesFromDataFrame(epi2)
  com <- sum(countOverlaps(epi1GR, epi2GR))
  return(c(epi1 = n1 - com, epi2 = n2 - com, Common = com))

}
boys.over <- lapply(seq_len(nrow(boys.comb)), function(i){
  getOverlaps(boys.comb[i, ]$method,
              boys.comb[i, ]$sample,
              res.inma4.boys.df, res.inma4.df)
})
boys.over.df <- Reduce(rbind, boys.over) %>%
  cbind(boys.comb[, c("method", "sample")], .) %>%
  tibble() %>%
  mutate(Boys = epi1, All = epi2)

boys.over.df %>%
  filter(!(All == 0 & Boys == 0 & Common == 0 )) %>%
  gather(Type, N, 5:7) %>%
  mutate(N = ifelse(N > 25, 25, N)) %>%
  ggplot(aes(x = N, fill = Type)) +
  geom_histogram(binwidth = 1) +
  facet_grid(method ~ .) +
  theme_bw()
```

In this plot, we show the epimutations detected only in boys, only in all and in common (boys and all) for those samples where at least one epimutation was detected. For barbosa, many individuals had 1 epimutation only detected in boys, while all the epimutations detected in the whole dataset were also detected in boys. 

In beta, we observed most individuals had only one epimutations that was detected in the whole and only boys datasets.

Finally, for mlm and manova we observed a high heterogeneity, with many individuals having epimutations specific of the boys or all datasets.

```{r}
epi_boys <- subset(res.inma4.boys.df, sample == "04_318_4" & method %in% c("beta", "barbosa"))

p1_boys <- plot_epimutations(epi_boys[1,], gset[, gset$Sex == "M"]) +
  ggtitle("Only boys")
p1_all <- plot_epimutations(epi_boys[1,], gset) +
  ggtitle("Whole dataset")
plot_grid(p1_boys, p1_all, ncol = 2)

```

In this region, we observe how some girls have lower methylation values than boys. Thus, the epimutation is only evident when comparing with boys but not with girls. 

## Girls


```{r }
load("../INMA4.epimutations.girls.Rdata")

res.inma4.girls.df <- make_res_df(res.inma4.girls.list) 
girls.sum.df <- make_sum_df(res.inma4.girls.df)
```



```{r}
girls.comb <- all.sum.df %>%
  ungroup() %>%
  mutate(all = n) %>%
  select(all, method, sample) %>% 
  right_join(girls.sum.df %>%
  ungroup() %>%
  mutate(girls = n) %>%
  select(girls, method, sample), by = c("method", "sample")) %>%
  mutate(Epimutations_diff = girls - all) 

ggplot(girls.comb, aes(x = Epimutations_diff)) +
  geom_histogram(binwidth = 1) +
  facet_grid(method ~ .) +
  theme_bw() +
  xlab("Difference in epimutations between girls and all")
```

For the four methods, the tendencies are the same than for boys. In barbosa and beta, most individuals had the same number of epimutations in the whole cohort or in girls only. Same individuals have few more or less epimutations while very few samples have much more epimutations in the girls only analysis.

For mlm and manova, we again had a very different number of epimutations depending on the dataset. 

```{r}
girls.over <- lapply(seq_len(nrow(girls.comb)), function(i){
  getOverlaps(girls.comb[i, ]$method,
              girls.comb[i, ]$sample,
              res.inma4.girls.df, res.inma4.df)
})
girls.over.df <- Reduce(rbind, girls.over) %>%
  cbind(girls.comb[, c("method", "sample")], .) %>%
  tibble() %>%
  mutate(Girls = epi1, All = epi2)

girls.over.df %>%
  filter(!(All == 0 & Girls == 0 & Common == 0 )) %>%
  gather(Type, N, 5:7) %>%
  mutate(Type = factor(Type, levels = c("All", "Girls", "Common"))) %>%
  mutate(N = ifelse(N > 25, 25, N)) %>%
  ggplot(aes(x = N, fill = Type)) +
  geom_histogram(binwidth = 1) +
  facet_grid(method ~ .) +
  theme_bw()
```

In this plot, we show the epimutations detected only in girls, only in all and in common (boys and all) for those samples where at least one epimutation was detected. In this case, we observed a high proportion of girls having 1 epimutation specific of the girls only dataset, as compared with boys. mlm and manova presented high heterogeneity, with many individuals having epimutations specific of the girls or all datasets.

```{r}
epi_girls <- subset(res.inma4.girls.df, sample == "04_59_4" & method %in% c("beta", "barbosa"))

p2_girls <- plot_epimutations(epi_girls[5,], gset[, gset$Sex == "F"]) +
  ggtitle("Only girls")
p2_all <- plot_epimutations(epi_girls[5,], gset) +
  ggtitle("Whole dataset")
plot_grid(p2_girls, p2_all, ncol = 2)
```

In this region, we observe how boys have higher methylation values than girls Thus, the epimutation is only evident when comparing with girls but not with the whole dataset. 

