---
title: "INMA_SAB0_epimutations"
author: "Carlos Ruiz"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Introduction

In this document, we will explore the epimutations detected by the different methods. We preselected the four methods that reported less epimutations: beta, mahdist, manova and mlm. We will also compare the difference between the two normalization procedures: independent and joint. We will focus on the epimutations detected in Esteller batch, the batch from the samples uesd as controls.

# Load data


```{r }
library(minfi)
library(tidyverse)
library(cowplot)
library(epimutacions)

load("../INMA0combined.normalizedComBat.autosomic.filterAnnotatedProbes.withNA.GenomicRatioSet.Rdata")
INMA_ind <- gset
load("../INMA_comb.normalizedComBat.autosomic.filterAnnotatedProbes.withNA.GenomicRatioSet.Rdata")
INMA_comb <- gset
load("../INMA0combined.epimutations.INMA0.duplicates.Rdata")
res_ind <- res
load("../INMA_comb.epimutations.INMA0.duplicates.Rdata")
res_comb <- res
load("../INMA.commonControlSamples.Rdata")

ind.Esteller <- INMA_ind[, INMA_ind$Batch == "Esteller"]
comb.Esteller <- INMA_comb[, INMA_comb$Batch == "Esteller"]


nMethod <- sapply(res_ind, nrow)
nMethod[sapply(nMethod, is.null)] <- 0

ind.res.df <- Reduce(rbind, lapply(res_ind, function(x) x[, 1:12])) %>%
  mutate(method = rep(names(res_ind), unlist(nMethod))) %>%
 left_join(colData(ind.Esteller) %>% 
              data.frame() %>% 
              select(Sample_Name, idnum, Batch) %>% 
              mutate(sample = Sample_Name), by = "sample") %>%
  filter(!is.na(Batch)) %>%
   mutate(Normalization = "Independent")


comb.res.df <- Reduce(rbind, lapply(res_comb, function(x) x[, 1:12])) %>%
  mutate(method = rep(names(res_comb), sapply(res_comb, nrow))) %>%
  left_join(colData(comb.Esteller) %>% 
              data.frame() %>% 
              select(Sample_Name, idnum, Batch) %>% 
              mutate(sample = Sample_Name), by = "sample") %>%
   filter(!is.na(Batch)) %>%
 mutate(Normalization = "Joint")

comSamps <- intersect(unique(ind.res.df$sample), unique(comb.res.df$sample))
  
sel.methods <- c("mlm", "beta", "manova", "mahdistmcd", "barbosa") 
names(sel.methods) <- sel.methods
all.res.df <- rbind(ind.res.df, comb.res.df) %>%
  filter(method %in% sel.methods) %>%
  filter(!is.na(sz)) %>%
  filter(Batch == "Esteller") %>%
  filter(pvalue < 0.05/40408 | is.na(pvalue)) %>%
  filter(sample %in% comSamps)

```


```{r}
plotEpimutationsSamples <- function(samp){
  df <- subset(all.res.df, sample == samp)
  gr <- makeGRangesFromDataFrame(df)
  com <- GenomicRanges::reduce(gr)
  
  methods.list <- lapply(sel.methods, function(x) subset(df, method == x))
  methods.list <- methods.list[sapply(methods.list, nrow) > 0]
  methods.GR <- lapply(methods.list, function(x){
    if (nrow(x) > 0){
      GenomicRanges::reduce(makeGRangesFromDataFrame(x))
    } else{
      return(NULL)
    }
  })
  
  ov.list <- lapply(methods.GR, findOverlaps, query = com)
  reg.tab <- table(unlist(lapply(ov.list, from)))
  
  reg.groups <- lapply(5:1, function(x) {
    names(reg.tab)[reg.tab == x]
  })
  
  fin.ov <- findOverlaps(com, gr)
  
  ## Plot By groups
  reg.plots <- lapply(reg.groups, function(x){
    if (length(x) == 0){
      return(NULL)
    }
    plots <- lapply(x, function(i){
      comGR <- com[as.numeric(i)]
      rows <- to(fin.ov)[from(fin.ov) == i]
      df.fil <- df[rows, ]
      
      df.fil.ind <- subset(df.fil, Normalization == "Independent")
      df.fil.joint <- subset(df.fil, Normalization == "Joint")
      ind <- plot_epimutations(df.fil[1, ], ind.Esteller[, c(samps, df.fil[1, ]$sample)]) +
        labs(title = "Independent Normalization",
             subtitle = ifelse(nrow(df.fil.ind) > 0,
                               paste("Detected by", paste(df.fil.ind$method, collapse = ", ")),
                               "Undetected")) +
        ylim(c(0, 1))
      comb <- plot_epimutations(df.fil[1, ], comb.Esteller[, c(samps, df.fil[1, ]$sample)]) +
        labs(title = "Joint Normalization",
             subtitle = ifelse(nrow(df.fil.joint) > 0,
                               paste("Detected by", paste(df.fil.joint$method, collapse = ", ")),
                               "Undetected")) +
        ylim(c(0, 1))
      plot_grid(ind, comb, ncol = 2)
    })
    plots
  })
  reg.plots
}
```

# SAB_C_0016

```{r}
plotEpimutationsSamples("SAB_C_0016")
```

# SAB_C_0017

```{r}
plotEpimutationsSamples("SAB_C_0017")
```

# SAB_C_0120

```{r}
plotEpimutationsSamples("SAB_C_0120")
```

# SAB_C_0211

```{r}
plotEpimutationsSamples("SAB_C_0211")
```

# SAB_C_0244

```{r}
plotEpimutationsSamples("SAB_C_0244")
```

# SAB_C_0340

```{r}
plotEpimutationsSamples("SAB_C_0340")
```

# SAB_C_0387

```{r}
plotEpimutationsSamples("SAB_C_0387")
```

# SAB_C_0636

```{r}
plotEpimutationsSamples("SAB_C_0636")
```