---
title: "INMA epimutations"
author: "Carlos Ruiz"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Introduction

In this document, we will analyze the epimutations identified in INMA. We used four methods to detected epimutations (beta, barbosa, mlm and manova). We compared each samples against the rest of the cohort (leave one out). We run the analysis including all samples, only boys or only girls. 


```{r}
library(tidyverse)
library(cowplot)
library(minfi)
library(robustbase)
library(epimutacions)

load("../INMA_comb.normalizedComBat.autosomic.filterAnnotatedProbes.withNA.GenomicRatioSet.Rdata")

```

# All samples

## Individual methods

We first evaluated the results of each method independently. In all the comparisons, we used a robust poisson model to account for individuals with an extremely large number of epimutations. 

```{r }
load("../INMA_comb.epimutations.allSamples.Rdata")

nMethod <- sapply(res.inma0.list, nrow)
nMethod[sapply(nMethod, is.null)] <- 0
methods <- names(res.inma0.list)
names(methods) <- methods
res.inma0.df <- Reduce(rbind, res.inma0.list) %>%
  mutate(method = rep(methods, unlist(nMethod))) %>%
 left_join(colData(gset) %>% 
              data.frame() %>% 
              select(Sample_Name, Sex, smkpreg, Batch) %>% 
              mutate(sample = Sample_Name), by = "sample")
all.sum.df <- res.inma0.df %>%
  group_by(method, sample, Sex, smkpreg, Batch) %>%
  summarize(n = sum(start != 0 & (is.na(pvalue) | pvalue < 0.05/40408))) %>%
  mutate(n_plot = ifelse(n > 10, "10+", as.character(n)),
         n_plot = factor(n_plot, levels = c(0:10, "10+")))

```

```{r}
all.sum.df %>%
  ggplot(aes(x = n_plot, fill = method)) +
  geom_bar(position=position_dodge()) +
  theme_bw()
summary(glmrob(n ~ method, all.sum.df, family = "poisson"))
```

As previously reported, barbosa is the method that detects less epimutations, followed by beta, mlm and manova. 

```{r}
all.sum.df %>%
  ggplot(aes(x = n_plot, fill = Sex)) +
  geom_bar(position=position_dodge()) +
  theme_bw() +
  facet_wrap(~ method, scales = "free")

lapply(methods, function(x){
  summary(glmrob(n ~ Sex, all.sum.df, subset = method == x, family = "poisson"))
})

```

We did not found differences between boys and girls.


```{r}
all.sum.df %>%
  ggplot(aes(x = n_plot, fill = Batch)) +
  geom_bar(position=position_dodge()) +
  theme_bw() +
  facet_wrap(~ method, scales = "free")

lapply(methods, function(x){
  summary(glmrob(n ~ Batch, all.sum.df, subset = method == x, family = "poisson"))
})

```

We found less epimutations in samples from MeDALL batch, possibly because it has a higher number of samples. 

```{r}
all.sum.df %>%
  filter(!is.na(smkpreg)) %>%
  ggplot(aes(x = n_plot, fill = smkpreg)) +
  geom_bar(position=position_dodge()) +
  theme_bw() +
  facet_wrap(~ method, scales = "free")

lapply(methods, function(x){
  summary(glmrob(n ~ smkpreg, all.sum.df, subset = method == x, family = "poisson"))
})
lapply(methods, function(x){
  summary(glmrob(n ~ smkpreg + Sex + Batch, all.sum.df, subset = method == x, family = "poisson"))
})

```

The association between the number of epimutations and maternal smoking was not consistent between the methods. Beta, barbosa and mlm found more epimutations when mother smoked during the whole pregnancy, although it was only significant for the beta. On the other hand, manova found less epimutations in children from smoker mothers. 



```{r}
lapply(methods, function(x){
  summary(glmrob(n ~ smkpreg, all.sum.df, subset = method == x & Batch == "Esteller", family = "poisson"))
})

lapply(methods[-c(1:2)], function(x){
  summary(glmrob(n ~ smkpreg, all.sum.df, subset = method == x & Batch == "MeDALL", family = "poisson"))
})
```

## Combined methods

Next, we considered regions detected by at least 2 methods. 

```{r}
all.sum.comb.df <-  res.inma0.df %>%
  group_by(epi_region_id, sample, Sex, smkpreg) %>%
  summarize(n = sum(start != 0 & (is.na(pvalue) | pvalue < 0.05/40408))) %>%
  group_by(sample, Sex, smkpreg) %>%
  summarize(n = sum(n >= 2)) %>%
  mutate(n_plot = ifelse(n > 10, "10+", as.character(n)),
         n_plot = factor(n_plot, levels = c(0:10, "10+")))
```


```{r}
all.sum.comb.df %>%
  ggplot(aes(x = n_plot, fill = Sex)) +
  geom_bar(position=position_dodge()) +
  theme_bw() 

summary(glmrob(n ~ Sex, all.sum.comb.df, family = "poisson"))

```

We again did not observe more epimutations in boys than in girls. 


```{r}
all.sum.comb.df %>%
  filter(!is.na(smkpreg)) %>%
  ggplot(aes(x = n_plot, fill = smkpreg)) +
  geom_bar(position=position_dodge()) +
  theme_bw() 
summary(glmrob(n ~ smkpreg, all.sum.comb.df, family = "poisson"))

summary(glmrob(n ~ smkpreg + Sex, all.sum.comb.df, family = "poisson"))
```

After combining the calls, we lost the association between smoking and epimutations.

# Stratification by sex

We evaluated the effect of running the epimutations in boys and girls independently.

## Boys

```{r }
load("../INMA_comb.epimutations.boys.Rdata")

nMethod.boys <- sapply(res.inma0.boys.list, nrow)
nMethod.boys[sapply(nMethod.boys, is.null)] <- 0
methods <- names(res.inma0.boys.list)
names(methods) <- methods
res.inma0.boys.df <- Reduce(rbind, res.inma0.boys.list) %>%
  mutate(method = rep(methods, unlist(nMethod.boys))) %>%
 left_join(colData(gset) %>% 
              data.frame() %>% 
              select(Sample_Name, Sex, smkpreg) %>% 
              mutate(sample = Sample_Name), by = "sample")
boys.sum.df <- res.inma0.boys.df %>%
  group_by(method, sample, Sex, smkpreg) %>%
  summarize(n = sum(start != 0 & (is.na(pvalue) | pvalue < 0.05/40408))) %>%
  mutate(n_plot = ifelse(n > 10, "10+", as.character(n)),
         n_plot = factor(n_plot, levels = c(0:10, "10+")))

```


```{r}
boys.comb <- all.sum.df %>%
  ungroup() %>%
  mutate(all = n) %>%
  select(all, method, sample) %>% 
  right_join(boys.sum.df %>%
  ungroup() %>%
  mutate(boys = n) %>%
  select(boys, method, sample), by = c("method", "sample")) %>%
  mutate(Epimutations_diff = boys - all) 

ggplot(boys.comb, aes(x = Epimutations_diff)) +
  geom_histogram(binwidth = 1) +
  facet_grid(method ~ .) +
  theme_bw() +
  xlab("Difference in epimutations between boys and all")
```

We observed very coincident number of epimutations between the whole cohort or using only boys when applying barbosa. Only for some cases, an epimutation was only detected when just analyzing boys. When using beta, most of the samples also had the same number of epimutations, although some individuals had very different number of epimutations between the whole cohort or the boys cohort. With manova, a higher proportion of samples had discrepancies, although the differences were of few epimutations. Finally, mlm had a low number of individuals with the same number of epimutations in the whole cohort or with only boys and many samples had a high number of epimutations differing between the analysis. 

```{r}
getOverlaps <- function(method, sample, epi_df1, epi_df2){
  
  epi1 <- epi_df1[epi_df1$method == method & epi_df1$sample == sample, ]
  epi1 <- subset(epi1, start != 0 & (is.na(pvalue) | pvalue < 0.05/40408))
  
  epi2 <- epi_df2[epi_df2$method == method & epi_df2$sample == sample, ]
  epi2 <- subset(epi2, start != 0 & (is.na(pvalue) | pvalue < 0.05/40408))

  
  n1 <- nrow(epi1)
  n2 <- nrow(epi2)
  
  if (n1 == 0 | n2 == 0){
    return(c(epi1 = n1, epi2 = n2, Common = 0))
  }
  epi1GR <- makeGRangesFromDataFrame(epi1)
  epi2GR <- makeGRangesFromDataFrame(epi2)
  com <- sum(countOverlaps(epi1GR, epi2GR))
  return(c(epi1 = n1 - com, epi2 = n2 - com, Common = com))

}
boys.over <- lapply(seq_len(nrow(boys.comb)), function(i){
  getOverlaps(boys.comb[i, ]$method,
              boys.comb[i, ]$sample,
              res.inma0.boys.df, res.inma0.df)
})
boys.over.df <- Reduce(rbind, boys.over) %>%
  cbind(boys.comb[, c("method", "sample")], .) %>%
  tibble() %>%
  mutate(Boys = epi1, All = epi2)

boys.over.df %>%
  filter(!(All == 0 & Boys == 0 & Common == 0 )) %>%
  gather(Type, N, 5:7) %>%
  mutate(N = ifelse(N > 25, 25, N)) %>%
  ggplot(aes(x = N, fill = Type)) +
  geom_histogram(binwidth = 1) +
  facet_grid(method ~ .) +
  theme_bw()
```

In this plot, we show the epimutations detected only in boys, only in all and in common (boys and all) for those samples where at least one epimutation was detected. For barbosa, most individuals had epimutations detected in boys and all datasets, so most individuals had 0 epimutations specific of boys or all datasets. For epimutations detected in one dataset, all of them were detected in the boys dataset.

In beta, we observed the same general trend, with even more individuals where all the epimutations detected were found in boys and all datasets. In this case, we also observed epimutations specific of the all dataset.

Finally, for mlm and manova we observed a high heterogeneity, with many individuals having epimutations specific of the boys or all datasets.

```{r}
epi_boys <- subset(res.inma0.boys.df, sample == "SAB_C_0223" & method %in% c("beta", "barbosa"))

p1_boys <- plot_epimutations(epi_boys[1,], gset[, gset$Sex == "M"]) +
  ggtitle("Only boys")
p1_all <- plot_epimutations(epi_boys[1,], gset) +
  ggtitle("Whole dataset")
plot_grid(p1_boys, p1_all, ncol = 2)

```

In this region, we observe how girls have higher methylation values than boys. Thus, the epimutation is only evident when comparing with boys but not with girls. 

## Girls


```{r }
load("../INMA_comb.epimutations.girls.Rdata")

nMethod.girls <- sapply(res.inma0.girls.list, nrow)
nMethod.girls[sapply(nMethod.girls, is.null)] <- 0
methods <- names(res.inma0.girls.list)
names(methods) <- methods
res.inma0.girls.df <- Reduce(rbind, res.inma0.girls.list) %>%
  mutate(method = rep(methods, unlist(nMethod.girls))) %>%
 left_join(colData(gset) %>% 
              data.frame() %>% 
              select(Sample_Name, Sex, smkpreg) %>% 
              mutate(sample = Sample_Name), by = "sample")
girls.sum.df <- res.inma0.girls.df %>%
  group_by(method, sample, Sex, smkpreg) %>%
  summarize(n = sum(start != 0 & (is.na(pvalue) | pvalue < 0.05/40408))) %>%
  mutate(n_plot = ifelse(n > 10, "10+", as.character(n)),
         n_plot = factor(n_plot, levels = c(0:10, "10+")))

```



```{r}
girls.comb <- all.sum.df %>%
  ungroup() %>%
  mutate(all = n) %>%
  select(all, method, sample) %>% 
  right_join(girls.sum.df %>%
  ungroup() %>%
  mutate(girls = n) %>%
  select(girls, method, sample), by = c("method", "sample")) %>%
  mutate(Epimutations_diff = girls - all) 

ggplot(girls.comb, aes(x = Epimutations_diff)) +
  geom_histogram(binwidth = 1) +
  facet_grid(method ~ .) +
  theme_bw() +
  xlab("Difference in epimutations between girls and all")
```

For barbosa, the results are very similar than for boys: most individuals have the same number of epimutations and some have more epimutations in girls than in all. For beta, we observe more coherent results between girls and all samples than when comparing boys with all samples. In this case, more smaples have more epimutations when considering all samples than only in girls, probably due to a lower statistical power. Finally, we observed again the more discordant results for manova and mlm.  

```{r}
girls.over <- lapply(seq_len(nrow(girls.comb)), function(i){
  getOverlaps(girls.comb[i, ]$method,
              girls.comb[i, ]$sample,
              res.inma0.girls.df, res.inma0.df)
})
girls.over.df <- Reduce(rbind, girls.over) %>%
  cbind(girls.comb[, c("method", "sample")], .) %>%
  tibble() %>%
  mutate(Girls = epi1, All = epi2)

girls.over.df %>%
  filter(!(All == 0 & Girls == 0 & Common == 0 )) %>%
  gather(Type, N, 5:7) %>%
  mutate(Type = factor(Type, levels = c("All", "Girls", "Common"))) %>%
  mutate(N = ifelse(N > 25, 25, N)) %>%
  ggplot(aes(x = N, fill = Type)) +
  geom_histogram(binwidth = 1) +
  facet_grid(method ~ .) +
  theme_bw()
```

In this plot, we show the epimutations detected only in girls, only in all and in common (boys and all) for those samples where at least one epimutation was detected. The trends are very similar than for boys: barbosa and beta presented few epimutations specific of girls or of all dataset, while mlm and manova presented high heterogeneity, with many individuals having epimutations specific of the girls or all datasets.

```{r}
epi_girls <- subset(res.inma0.girls.df, sample == "SAB_C_0048" & method %in% c("beta", "barbosa"))

p2_girls <- plot_epimutations(epi_girls[1,], gset[, gset$Sex == "F"]) +
  ggtitle("Only girls")
p2_all <- plot_epimutations(epi_girls[1,], gset) +
  ggtitle("Whole dataset")
plot_grid(p2_girls, p2_all, ncol = 2)
```

In this region, we observe how boys have higher methylation values than girls Thus, the epimutation is only evident when comparing with girls but not with the whole dataset. 



# Stratification by bacth

We evaluated the effect of running the epimutations in each batch independently.

## Esteller

```{r }
load("../INMA_comb.epimutations.esteller.Rdata")

nMethod.esteller <- sapply(res.inma0.esteller.list, nrow)
nMethod.esteller[sapply(nMethod.esteller, is.null)] <- 0
methods <- names(res.inma0.esteller.list)
names(methods) <- methods
res.inma0.esteller.df <- Reduce(rbind, res.inma0.esteller.list) %>%
  mutate(method = rep(methods, unlist(nMethod.esteller))) %>%
 left_join(colData(gset) %>% 
              data.frame() %>% 
              select(Sample_Name, Sex, smkpreg) %>% 
              mutate(sample = Sample_Name), by = "sample")
esteller.sum.df <- res.inma0.esteller.df %>%
  group_by(method, sample, Sex, smkpreg) %>%
  summarize(n = sum(start != 0 & (is.na(pvalue) | pvalue < 0.05/40408))) %>%
  mutate(n_plot = ifelse(n > 10, "10+", as.character(n)),
         n_plot = factor(n_plot, levels = c(0:10, "10+")))

```


```{r}
esteller.comb <- all.sum.df %>%
  ungroup() %>%
  mutate(all = n) %>%
  select(all, method, sample) %>% 
  right_join(esteller.sum.df %>%
  ungroup() %>%
  mutate(esteller = n) %>%
  select(esteller, method, sample), by = c("method", "sample")) %>%
  mutate(Epimutations_diff = esteller - all) 

ggplot(esteller.comb, aes(x = Epimutations_diff)) +
  geom_histogram(binwidth = 1) +
  facet_grid(method ~ .) +
  theme_bw() +
  xlab("Difference in epimutations between Esteller and all")
```

The four methods reported a similar number of epimutations when considering Esteller samples or when considering the full dataset. Few samples in beta had much more epimutations in Esteller, while some samples have one or two epimutations different in Esteller compared to the whole dataset. 


```{r}
esteller.over <- lapply(seq_len(nrow(esteller.comb)), function(i){
  getOverlaps(esteller.comb[i, ]$method,
              esteller.comb[i, ]$sample,
              res.inma0.esteller.df, res.inma0.df)
})
esteller.over.df <- Reduce(rbind, esteller.over) %>%
  cbind(esteller.comb[, c("method", "sample")], .) %>%
  tibble() %>%
  mutate(Esteller = epi1, All = epi2)

esteller.over.df %>%
  filter(!(All == 0 & Esteller == 0 & Common == 0 )) %>%
  gather(Type, N, 5:7) %>%
  mutate(Type = factor(Type, levels = c("All", "Esteller", "Common"))) %>%
  mutate(N = ifelse(N > 25, 25, N)) %>%
  ggplot(aes(x = N, fill = Type)) +
  geom_histogram(binwidth = 1) +
  facet_grid(method ~ .) +
  theme_bw()
```

However, when we compare in the epimutations detected are the same in the whole dataset or using just Esteller, we observe some differences between the methods. Thus, beta and barbosa are robust and most of the epimutations are identified in the whole dataset and in Esteller. On the ohter hand, manova and mlm have higher heterogeneity and detected some epimutations only in one of the datasets. 

```{r}
esteller.sum.df %>%
  filter(!is.na(smkpreg)) %>%
  ggplot(aes(x = n_plot, fill = smkpreg)) +
  geom_bar(position=position_dodge()) +
  theme_bw() +
  facet_wrap(~ method, scales = "free")

lapply(methods, function(x){
  summary(glmrob(n ~ smkpreg, esteller.sum.df, subset = method == x, family = "poisson"))
})
lapply(methods, function(x){
  summary(glmrob(n ~ smkpreg + Sex , esteller.sum.df, subset = method == x, family = "poisson"))
})

```

When considering only Esteller samples, we observed that children whose mother smoked throughout the whole pregnancy had more epimutations, although this is only significant for mlm.

## MeDALL


```{r }
load("../INMA_comb.epimutations.medall.Rdata")

nMethod.medall <- sapply(res.inma0.medall.list, nrow)
nMethod.medall[sapply(nMethod.medall, is.null)] <- 0
methods <- names(res.inma0.medall.list)
names(methods) <- methods
res.inma0.medall.df <- Reduce(rbind, res.inma0.medall.list) %>%
  mutate(method = rep(methods, unlist(nMethod.medall))) %>%
 left_join(colData(gset) %>% 
              data.frame() %>% 
              select(Sample_Name, Sex, smkpreg) %>% 
              mutate(sample = Sample_Name), by = "sample")
medall.sum.df <- res.inma0.medall.df %>%
  group_by(method, sample, Sex, smkpreg) %>%
  summarize(n = sum(start != 0 & (is.na(pvalue) | pvalue < 0.05/40408))) %>%
  mutate(n_plot = ifelse(n > 10, "10+", as.character(n)),
         n_plot = factor(n_plot, levels = c(0:10, "10+")))

```



```{r}
medall.comb <- all.sum.df %>%
  ungroup() %>%
  mutate(all = n) %>%
  select(all, method, sample) %>% 
  right_join(medall.sum.df %>%
  ungroup() %>%
  mutate(medall = n) %>%
  select(medall, method, sample), by = c("method", "sample")) %>%
  mutate(Epimutations_diff = medall - all) 

ggplot(medall.comb, aes(x = Epimutations_diff)) +
  geom_histogram(binwidth = 1) +
  facet_grid(method ~ .) +
  theme_bw() +
  xlab("Difference in epimutations between MeDALL and all")
```

The four methods reported a similar number of epimutations when considering MeDALL samples or when considering the full dataset. Few samples in beta had much less epimutations in MeDALL, while some samples have one or two epimutations different in Esteller compared to the whole dataset. 


```{r}
medall.over <- lapply(seq_len(nrow(medall.comb)), function(i){
  getOverlaps(medall.comb[i, ]$method,
              medall.comb[i, ]$sample,
              res.inma0.medall.df, res.inma0.df)
})
medall.over.df <- Reduce(rbind, medall.over) %>%
  cbind(medall.comb[, c("method", "sample")], .) %>%
  tibble() %>%
  mutate(MeDALL = epi1, All = epi2)

medall.over.df %>%
  filter(!(All == 0 & MeDALL == 0 & Common == 0 )) %>%
  gather(Type, N, 5:7) %>%
  mutate(Type = factor(Type, levels = c("All", "MeDALL", "Common"))) %>%
  mutate(N = ifelse(N > 25, 25, N)) %>%
  ggplot(aes(x = N, fill = Type)) +
  geom_histogram(binwidth = 1) +
  facet_grid(method ~ .) +
  theme_bw()
```

Most samples had no epimutations for beta or barbosa, either in the whole dataset or when considering only MeDALL. As a result, the bars of the histogram are much lower. Again, we observe higher heterogeneity with mlm or manova and more individuals have very high number of epimutations. 


```{r}
medall.sum.df %>%
  filter(!is.na(smkpreg)) %>%
  ggplot(aes(x = n_plot, fill = smkpreg)) +
  geom_bar(position=position_dodge()) +
  theme_bw() +
  facet_wrap(~ method, scales = "free")

lapply(methods[-2], function(x){
  summary(glmrob(n ~ smkpreg, medall.sum.df, subset = method == x, family = "poisson"))
})
lapply(methods[-2], function(x){
  summary(glmrob(n ~ smkpreg + Sex , medall.sum.df, subset = method == x, family = "poisson"))
})
```

Again, we observe more epimutation in children whose mother smoked throughout the whole pregnancy, but the association is not significant. 
