---
title: "INMA 0 years QC"
author: "Carlos Ruiz"
date: "18/2/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Introduction

The goal of this document is to show the results of the QC of the INMA 0 years data. INMA data was processed in two batches (MeDALL and Esteller). We tried two processing approaches:
1. Normalize each batch separately and merge the normalized results (independent normalization).
2. Normalize both batches jointly (joint normalization).

```{r }
library(minfi)
library(meffil)
library(tidyverse)
library(cowplot)

load("../INMA0combined.normalizedComBat.autosomic.filterAnnotatedProbes.GenomicRatioSet.Rdata")
INMA_ind <- gset
load("../INMA_comb.normalizedComBat.autosomic.filterAnnotatedProbes.GenomicRatioSet.Rdata")
INMA_comb <- gset
load("../INMA0combined.epimutations.INMA0.duplicates.Rdata")
res_ind <- res
load("../INMA_comb.epimutations.INMA0.duplicates.Rdata")
res_comb <- res
```

# DNA methylation normalization
## Independent normalization

The results of this section refer to the independent normalization. 

### Combined dataset

We will first evaluate the correlation of different variables with the principal components of methylation of the combined dataset.

```{r}
ind.pcs <- meffil.methylation.pcs(getBeta(INMA_ind), probe.range = 40000)
ind.pcs.df <- ind.pcs %>%
  data.frame() %>%
  select(PC1, PC2, PC3, PC4) %>%
  mutate(Sample_Name = rownames(.)) %>%
  left_join(colData(INMA_ind) %>% data.frame() %>% select(Sample_Name, Sex, dup, Batch, idnum), by = "Sample_Name") %>%
  mutate(Sample = ifelse(dup, idnum, "Reference"))
```


```{r}
makePlots <- function(df, batch){
  p1 <- ggplot(df, aes_string(x = "PC1", y = "PC2", color = batch)) +
    geom_point() + theme_bw()
  p2 <- ggplot(df, aes_string(x = "PC3", y = "PC4", color = batch)) +
    geom_point() + theme_bw()
  plot_grid(p1, p2, nrow = 1)
}
```

#### Batch

```{r}
makePlots(ind.pcs.df, "Batch")
```

The samples are clearly separated by batch. 

#### Replicates

We will compare how different the same sample processed in a different batch. 

```{r}
ggplot(ind.pcs.df, aes_string(x = "PC1", y = "PC2", shape = "Batch")) +
    geom_point() + theme_bw() +
  geom_line(data = subset(ind.pcs.df, dup), aes(group = Sample, color = Sample))
```

Each replicate falls inside their batch, so the same sample has a different methylation pattern depending on its processing batch.

#### Sex

```{r}
makePlots(ind.pcs.df, "Sex")
```


We could not observed a correlation of sex with the principal components. 


### Esteller + replicates

In order to test how the epimutations method work when samples are processed in different batches, we will take samples from Esteller batch and the replicates from MeDALL. We will also test whether the batch effects are visible and whether sex has a big influence on overall methylation.

```{r}
ind.Esteller <- INMA_ind[, INMA_ind$Batch == "Esteller" | INMA_ind$dup]
ind.esteller.pcs <- meffil.methylation.pcs(getBeta(ind.Esteller), probe.range = 40000)
ind.esteller.pcs.df <- ind.esteller.pcs %>%
  data.frame() %>%
  select(PC1, PC2, PC3, PC4) %>%
  mutate(Sample_Name = rownames(.)) %>%
  left_join(colData(ind.Esteller) %>% data.frame() %>% select(Sample_Name, Sex, dup, Batch, idnum), by = "Sample_Name") %>%
  mutate(Sample = ifelse(dup, idnum, "Reference"))
```


#### Batch

```{r}
makePlots(ind.esteller.pcs.df, "Batch")
```

Differences due to batch are now only evident in the fourth PC.

#### Replicates

We will compare how different the same sample processed in a different batch. 

```{r}
ggplot(ind.esteller.pcs.df, aes_string(x = "PC1", y = "PC2")) +
    geom_point() + theme_bw() +
  geom_line(data = subset(ind.esteller.pcs.df, dup), aes(group = Sample, color = Sample))
```

Although samples are closer in the PCs, we still see some differences.

#### Sex

```{r}
makePlots(ind.esteller.pcs.df, "Sex")
```

We do not observe an effect of sex for Esteller samples only. 

## Joint normalization

The results of this section refer to the joint normalization. 

### Combined dataset

We will first evaluate the correlation of different variables with the principal components of methylation of the combined dataset.

```{r}
comb.pcs <- meffil.methylation.pcs(getBeta(INMA_comb), probe.range = 40000)
comb.pcs.df <- comb.pcs %>%
  data.frame() %>%
  select(PC1, PC2, PC3, PC4) %>%
  mutate(Sample_Name = rownames(.)) %>%
  left_join(colData(gset) %>% data.frame() %>% select(Sample_Name, Sex, dup, Batch, idnum), by = "Sample_Name") %>%
  mutate(Sample = ifelse(dup, idnum, "Reference"))
```

#### Batch

```{r}
makePlots(comb.pcs.df, "Batch")
```

There are no differences due to batch. 

#### Replicates

We will compare how different the same sample processed in a different batch. 

```{r}
ggplot(comb.pcs.df, aes_string(x = "PC1", y = "PC2", shape = "Batch")) +
    geom_point() + theme_bw() +
  geom_line(data = subset(comb.pcs.df, dup), aes(group = Sample, color = Sample))
```

Although at a general level there are no consistent differences between batch, replicate samples are separated in the PCA.

#### Sex

```{r}
makePlots(comb.pcs.df, "Sex")
```

We could not observed a correlation of sex with the principal components. 

### Esteller + replicates

In order to test how the epimutations method work when samples are processed in different batches, we will take samples from Esteller batch and the replicates from MeDALL. We will also test whether the batch effects are visible and whether sex has a big influence on overall methylation.

```{r}
comb.gset.Esteller <- INMA_comb[, INMA_comb$Batch == "Esteller" | INMA_comb$dup]
comb.esteller.pcs <- meffil.methylation.pcs(getBeta(comb.gset.Esteller), probe.range = 40000)
comb.esteller.pcs.df <- comb.esteller.pcs %>%
  data.frame() %>%
  select(PC1, PC2, PC3, PC4) %>%
  mutate(Sample_Name = rownames(.)) %>%
  left_join(colData(comb.gset.Esteller) %>% data.frame() %>% select(Sample_Name, Sex, dup, Batch, idnum), by = "Sample_Name") %>%
  mutate(Sample = ifelse(dup, idnum, "Reference"))
```


#### Batch

```{r}
makePlots(comb.esteller.pcs.df, "Batch")
```

Some MeDALL samples are outliers in the PC3. The remaining are inside the general distribution in all the PCs.

#### Replicates

We will compare how different the same sample processed in a different batch. 

```{r}
ggplot(comb.esteller.pcs.df, aes_string(x = "PC1", y = "PC2")) +
    geom_point() + theme_bw() +
  geom_line(data = subset(comb.esteller.pcs.df, dup), aes(group = Sample, color = Sample))
```

We still see some differences between the replicates in the PCA.

#### Sex

```{r}
makePlots(comb.esteller.pcs.df, "Sex")
```

We do not observe an effect of sex for Esteller samples only. 

# Epimutations detection

We will compare the epimutations detected by each replicate, either in the independent or the joint normalization. We will consider all the methods included in the `epimutacions` package.

```{r}
nMethod <- sapply(res_ind, nrow)
nMethod[sapply(nMethod, is.null)] <- 0
ind.ids <- ind.Esteller$idnum[duplicated(ind.Esteller$idnum)]

ind.res.df <- Reduce(rbind, res_ind) %>%
  mutate(method = rep(names(res_ind), unlist(nMethod))) %>%
 full_join(colData(ind.Esteller) %>% 
              data.frame() %>% 
              filter(idnum %in% ind.ids) %>% 
              select(Sample_Name, idnum, Batch) %>% 
              mutate(sample = Sample_Name), by = "sample") %>%
  filter(!is.na(Batch)) %>%
   mutate(Normalization = "Independent")


comb.ids <- comb.gset.Esteller$idnum[duplicated(comb.gset.Esteller$idnum)]
comb.res.df <- Reduce(rbind, res_comb) %>%
  mutate(method = rep(names(res_comb), sapply(res_comb, nrow))) %>%
  full_join(colData(comb.gset.Esteller) %>% 
              data.frame() %>% 
              filter(idnum %in% comb.ids) %>% 
              select(Sample_Name, idnum, Batch) %>% 
              mutate(sample = Sample_Name), by = "sample") %>%
   filter(!is.na(Batch)) %>%
 mutate(Normalization = "Joint")
  
all.res.df <- rbind(ind.res.df, comb.res.df) %>%
  filter(method != "barbosa")
p1 <- all.res.df  %>%
  group_by(Batch, idnum, method, Normalization) %>%
  summarize(n = sum(is.na(pvalue) | adj_pvalue < 0.05)) %>%
   mutate(Replicate = factor(idnum)) %>%
  ggplot(aes(x = Batch, y = n, color = Replicate, group = Replicate)) +
  geom_point() + geom_line() + theme_bw() + 
  ggtitle("Adjusted p-value < 0.05") +
  facet_grid(method ~ Normalization, scales = "free_y")

p2 <- all.res.df %>%
 filter(idnum != "339") %>%
  group_by(Batch, idnum, method, Normalization) %>%
  summarize(n = sum(is.na(pvalue) | adj_pvalue < 0.05)) %>%
   mutate(Replicate = factor(idnum)) %>%
  ggplot(aes(x = Batch, y = n, color = Replicate, group = Replicate)) +
  geom_point() + geom_line() + theme_bw() + 
  ggtitle("Adjusted p-value < 0.05, outlier removed") +
  facet_grid(method ~ Normalization, scales = "free_y")
plot_grid(p1, p2, ncol = 2 )


```

When datasets are independently normalized, we observe a higher number of epimutations in the replicate from the external batch (MeDALL) than in the replicate form the reference batch (Esteller). This effect is not so obvious when both datasets are normalized jointly. In general, there is high heterogeneity in the number of epimutations detected.

```{r}
p3 <- all.res.df %>%
 filter(method %in% c("manova", "mlm")) %>%
  group_by(Batch, idnum, method, Normalization) %>%
  summarize(n = sum(is.na(pvalue) | pvalue < 0.05/40408)) %>%
   mutate(Replicate = factor(idnum)) %>%
  ggplot(aes(x = Batch, y = n, color = Replicate, group = Replicate)) +
  geom_point() + geom_line() + theme_bw() + 
  ggtitle("p-value < 0.05/40408") +
  facet_grid(method ~ Normalization, scales = "free_y")


p4 <- all.res.df %>%
 filter(method %in% c("manova", "mlm") & idnum != "339") %>%
  group_by(Batch, idnum, method, Normalization) %>%
  summarize(n = sum(is.na(pvalue) | pvalue < 0.05/40408)) %>%
   mutate(Replicate = factor(idnum)) %>%
  ggplot(aes(x = Batch, y = n, color = Replicate, group = Replicate)) +
  geom_point() + geom_line() + theme_bw() + 
  ggtitle("p-value < 0.05/40408") +
  facet_grid(method ~ Normalization, scales = "free_y")

plot_grid(p3, p4, ncol = 2 )

```

The tendency is the same when considering all possible regions as the number of effective tests. 


```{r}
ids <- unique(all.res.df$idnum)
methods <- unique(all.res.df$method)
norms <- unique(all.res.df$Normalization)

comb <- data.frame(expand_grid(ids, methods, norms))
overlaps <- sapply(seq_len(nrow(comb)), function(x) {
  sub <- subset(all.res.df, idnum == comb[x, 1] & method == comb[x, 2] & Normalization == comb[x, 3] &
                  (pvalue < 0.05/40408 | is.na(pvalue))  )
  
  if (sum(sub$Batch == "MeDALL") == 0 | sum(sub$Batch == "Esteller") == 0){
    return(c(Common = 0, medall = 0, esteller = 0))
  }
  if (nrow(sub) == 0){
    return(c(Common = 0, medall = 0, esteller = 0))
  }
  
  medall <- makeGRangesFromDataFrame(subset(sub, Batch == "MeDALL"))
  est <- makeGRangesFromDataFrame(subset(sub, Batch == "Esteller"))
  com <- sum(countOverlaps(medall, est))
  c(Common = com, medall = length(medall) - com, esteller = length(est) - com)
})
overlaps.df <- overlaps %>%
  t() %>%
  cbind(comb) %>%
  tibble() %>%
  filter(!(Common == 0 & medall == 0 & esteller == 0)) %>%
  filter(!(ids %in% c("204", "484", "655", "339")))

overlaps.df %>%
  gather("Batch", "Regions", 1:3) %>%
  ggplot(aes(x = norms, y = Regions, fill = Batch)) +
    geom_bar(stat="identity") +
  theme_bw() + 
  facet_grid(methods ~ ids, scales = "free_y") +
   theme(axis.text.x = element_text(angle=90, vjust=0.5))
```

We observed a high variability in the number of epimutations detected between the methods and the individuals. Nonetheless, some general trends were visible. When normalizing the datasets independently, a higher number of epimutations is detected in the replicate from the external batch, as suggested by the previous plot. However, when datasets are normalized jointly, we observed a higher number  of epimutations in the replicate from the reference batch.

```{r}
overlaps.df %>%
  gather("Batch", "Regions", 1:3) %>%
  group_by(norms, methods, ids) %>%
  mutate(Prop = Regions/sum(Regions)) %>%
  ggplot(aes(x = norms, y = Prop, fill = Batch)) +
    geom_bar(stat="identity") +
  theme_bw() + 
  facet_grid(methods ~ ids, scales = "free_y") +
   theme(axis.text.x = element_text(angle=90, vjust=0.5))
```

When transforming the results to proportions, the main trends remained: independent normalization returned a higher proportion of epimutations in the replicate from the external batch and joint normalization in the replicate from the reference batch. mahdistmcd was the method that reported more consistent epimutations in all conditions. This method also returned a lower number of epimutations, suggesting that is more restrictive. Interistingly, this method had a big increase in the proportion of common epimutations when using joint normalization as opposed to independent.

Finally, we compared the epimutations detected in the Esteller batch (same batch than used for reference) when normalizing the cohort independently or jointly. In this case, we are comparing the epimutations detected in the same biological sample, we change whether the case and control samples were normalized alone or with the other cohort. 

```{r}
combids <- data.frame(expand_grid(ids, methods))
esteller.df <- subset(all.res.df, Batch == "Esteller")
overlapsEsteller <- sapply(seq_len(nrow(combids)), function(x) {
  sub <- subset(esteller.df, idnum == combids[x, 1] & method == combids[x, 2] &
                  (pvalue < 0.05/40408 | is.na(pvalue))  )
  
  if (sum(sub$Normalization == "Independent") == 0 | sum(sub$Normalization == "Joint") == 0){
    return(c(common = 0, independent = 0, joint = 0))
  }
  if (nrow(sub) == 0){
    return(c(common = 0, independent = 0, joint = 0))
  }
  
  indep <- makeGRangesFromDataFrame(subset(sub, Normalization == "Independent"))
  joint <- makeGRangesFromDataFrame(subset(sub, Normalization == "Joint"))
  com <- sum(countOverlaps(indep, joint))
  c(common = com, independent = length(indep) - com, joint = length(joint) - com)
})
overlapsEsteller.df <- overlapsEsteller %>%
  t() %>%
  cbind(combids) %>%
  tibble() %>%
  filter(!(common == 0 & independent == 0 & joint == 0)) %>%
  filter(!(ids %in% c("204", "484", "655", "339")))


overlapsEsteller.df %>%
  gather("Normalization", "Regions", 1:3) %>%
  ggplot(aes(x = as.factor(ids), y = Regions, fill = Normalization)) +
    geom_bar(stat="identity") +
  theme_bw() + 
  facet_grid(methods ~ ., scales = "free_y") +
   theme(axis.text.x = element_text(angle=90, vjust=0.5))
```

We observed a very strong effect of the normalization on the epimutations detected: the same samples can lead to different detection of epimutations, depending if they are normalized alone or together with other samples. We do not observe big differences between the methods, with differences, between the individuals. 




